---
title: "Chiral Graviton Modes in Fractional Quantum Hall Liquids"
subtitle: "Quantum Geometry, Spectral Sum Rules, and Polarization Selection Rules in R"
author: "John Akwei, Senior Data Scientist"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    df_print: kable
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = 'center', cache = TRUE)
```

```{r libraries, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(gridExtra)

theme_set(theme_minimal())
```

<br />

# Abstract

Chiral graviton modes (CGMs) are the long-wavelength, spin-2 limit of the
Girvin–MacDonald–Platzman (GMP) magnetoroton in fractional quantum Hall (FQH)
liquids. They are the quanta of fluctuations of Haldane's *guiding-centre
metric* — an internal, emergent geometric degree of freedom of the electron
fluid, not of spacetime. This document (i) corrects and tightens a chat-export
summary of the Nanjing University experiments, (ii) implements the exact
topological bookkeeping that governs graviton chirality (Wen–Zee shift,
guiding-centre spin, Haldane's bound, the Golkar–Nguyen–Son spectral sum
rules), and (iii) maps the genuine technical dependencies — as opposed to
structural analogies — between this material and the existing repository
documents `Gravitation.md`, `Gravitons_Gluons_Branes.md`,
`Quantum_Chromodynamics_in_R.Rmd`, `Fracton_Codes_in_R.Rmd`, and
`QuasiParticles_in_R.Rmd`.

The central quantitative result is that graviton chirality is **not** an
observation to be reported — it is a **theorem** with a topological input. A
non-negativity argument applied to two spectral densities forces
$S_4 \ge |\mathcal{S}-1|/8$, and equality holds *if and only if* one circular
polarization channel is empty. The experiment is a test of how close a real
Coulomb ground state sits to that bound.

<br />

# Epistemic flag convention

Carried over from `Gravitons_Gluons_Branes.md` for consistency across the
repository.

- **[Established]** — experimental result or theorem; no serious dispute.
- **[Standard theory]** — mainstream theoretical result within its framework.
- **[Model]** — true within a specific trial wavefunction or effective model;
  not expected to hold for generic Hamiltonians.
- **[Conjectural]** — proposed but unproven.
- **[Corrected]** — the source chat transcript stated this incorrectly.
- **[Convention]** — sign or normalization depends on a stated choice; check
  against the cited paper before quoting a number.

<br />

# What was actually measured

Two distinct experiments are conflated in the source transcript. They are two
years apart and report different things. **[Corrected]**

| | 2024 result | 2026 result |
|---|---|---|
| Reference | Liang, Liu, Yang, Huang, Wurstbauer, Dean, West, Pfeiffer, Du, Pinczuk, *Nature* **628**, 78–83 (2024) | Yang, Wang, Lu, Liu, Baldwin, West, Pfeiffer, Yang, Balram, Du, *Nature Physics* (2026), doi:10.1038/s41567-026-03338-9 |
| Claim | First observation of a chiral spin-2 long-wavelength magnetoroton | Observation of *multiple* gravitons, incl. a high-energy mode |
| Fillings | $\nu = 1/3$ (and related) | $\nu = 1/4$, $2/7$, $2/9$ |
| Interpretation | Supports Haldane's geometric description | Supports the **parton** construction of FQH states |
| Technique | Circularly polarized resonant inelastic light scattering (CP-RILS) | Same architecture |

## Corrections to the source transcript

1. **"Chirality of $\pm 2$" was not observed.** **[Corrected]** The physical
   content is the opposite: at $\nu = 1/3$ the mode appears in *one* circular
   polarization channel only, corresponding to angular-momentum transfer
   $-2$. A fundamental graviton has *both* helicities $\pm 2$; the FQH mode is
   *maximally chiral*, and that asymmetry is the signature. For particle–hole
   conjugate states such as $\nu = 2/3$ the sign reverses to $+2$.
   **[Established]**

2. **"Behave precisely like the hypothetical fundamental graviton" is an
   overclaim.** **[Corrected]** What is shared is a single quantum number:
   charge $\pm 2$ under the $SO(2)$ rotation group of the plane. See
   §"What the CGM is not".

3. **"First experimental evidence ... published in Nature" dated to 2026.**
   **[Corrected]** The first evidence is the March 2024 *Nature* paper. The
   2026 *Nature Physics* paper is a follow-up on partons.

4. **"Benchtop simulator to test the complex tensor mathematics behind quantum
   gravity."** **[Corrected]** It tests Haldane's geometric theory of the FQH
   effect. It supplies no information about graviton self-coupling, the
   Einstein–Hilbert action, or any UV completion of gravity.

5. **"Parton" is a false cognate with QCD.** **[Corrected]** In the FQH
   context (Jain 1989) partons are unphysical fractionally charged fermions
   into which an electron is formally decomposed and then reassembled. They
   are *not* the Feynman–Bjorken partons of
   `Quantum_Chromodynamics_in_R.Rmd`. Structural analogy only.

6. **Resonance is not incidental.** **[Established]** The "R" in CP-RILS
   matters: the light-scattering cross-section is only measurable when the
   laser is tuned near the fundamental optical transition of the GaAs quantum
   well. The transcript treats RILS as generic spectroscopy.

7. **A subtlety the transcript skips.** The GMP density-wave mode has
   *vanishing* spectral weight in the density channel as $q \to 0$. The
   reason a $q \to 0$ measurement is possible at all is the Golkar–Nguyen–Son
   result that the intensity in the **spin-2 (stress-tensor) channel** does
   not vanish at $q = 0$. Without that, the experiment would have no signal.
   **[Standard theory]**

<br />

# Quantum geometry: why "spin-2" at all

Restricting to the lowest Landau level leaves only the guiding-centre
coordinates $R^a$, which do not commute:

$$[R^a, R^b] = -i \ell_B^2 \epsilon^{ab}, \qquad \ell_B = \sqrt{\hbar / eB}.$$

Haldane's observation is that the FQH ground state must choose a **unimodular
metric** $g_{ab}$ ($\det g = 1$) to define the shape of the correlation hole
around each guiding centre. A unimodular symmetric $2\times 2$ matrix has two
free parameters, which package into a single complex field carrying charge
$\pm 2$ under rotations. That is the entire origin of "spin-2." **[Standard
theory]**

Two consequences follow immediately, and both are worth stating because they
close off the loose analogy in the source transcript:

- The metric is **internal** to the electron fluid, not a fluctuation of the
  ambient spacetime metric. The GaAs lattice is unaffected.
- "Spin-2" here means charge $2$ under $SO(2)$, the rotation group of a plane.
  A four-dimensional graviton lives in a spin-2 representation of the Lorentz
  group $SO(3,1)$, which is a strictly stronger statement. **[Established]**

<br />

# Shift, guiding-centre spin, and Haldane's bound

The topological input is the Wen–Zee shift $\mathcal{S}$, defined on the
sphere by $N_\phi = \nu^{-1} N - \mathcal{S}$. The guiding-centre spin is
$\bar{s} = (\mathcal{S} - 1)/2$.

Haldane's inequality on the $q^4$ coefficient of the projected static
structure factor is

$$S_4 \;\ge\; \frac{|\mathcal{S} - 1|}{8},$$

proven as a sum rule by Golkar, Nguyen and Son (*JHEP* **2016**, 21).
**[Established]**

```{r shift_table}
# Shift formulas -------------------------------------------------------------
# Laughlin  nu = 1/m                 : S = m
# Jain forward-flux nu = n/(2pn + 1) : S = n + 2p
# Jain reverse-flux nu = n/(2pn - 1) : S = 2p - n
# [Standard theory] -- these are the standard LLL conventions.

fqh_states <- bind_rows(
  data.frame(family = "IQHE",     label = "1",    nu = 1,        S = 1),
  data.frame(family = "Laughlin", label = "1/3",  nu = 1/3,      S = 3),
  data.frame(family = "Laughlin", label = "1/5",  nu = 1/5,      S = 5),
  data.frame(family = "Laughlin", label = "1/7",  nu = 1/7,      S = 7),
  data.frame(family = "Jain +",   label = "2/5",  nu = 2/5,      S = 4),
  data.frame(family = "Jain +",   label = "3/7",  nu = 3/7,      S = 5),
  data.frame(family = "Jain +",   label = "2/9",  nu = 2/9,      S = 6),
  data.frame(family = "Jain -",   label = "2/3",  nu = 2/3,      S = 0),
  data.frame(family = "Jain -",   label = "3/5",  nu = 3/5,      S = -1),
  data.frame(family = "Jain -",   label = "2/7",  nu = 2/7,      S = 2)
) %>%
  mutate(
    s_bar        = (S - 1) / 2,
    S4_bound     = abs(S - 1) / 8,
    chirality    = ifelse(S == 1, 0, ifelse(S > 1, -2, +2)),
    chirality_lb = factor(chirality, levels = c(-2, 0, 2),
                          labels = c("-2", "none", "+2"))
  )

knitr::kable(
  fqh_states %>% select(family, label, S, s_bar, S4_bound, chirality),
  col.names = c("Family", "nu", "Shift S", "s-bar", "Haldane bound on S4",
                "Predicted chirality"),
  digits = 4,
  caption = "Topological bookkeeping for the chiral graviton mode."
)
```

**Consistency checks.** Two independent facts must fall out, and they do:

- $\nu = 1$: $\mathcal{S} = 1 \Rightarrow \bar{s} = 0 \Rightarrow$ bound is
  $0$. The integer quantum Hall state has no graviton. This is the control
  model for the whole framework.
- $\nu = 2/3$: $\mathcal{S} = 0 \Rightarrow \mathcal{S} - 1 < 0
  \Rightarrow$ chirality $+2$, opposite to $\nu = 1/3$. This reproduces the
  known particle–hole reversal without any additional input. **[Established]**

## Laughlin saturation

For the Laughlin wavefunction at $\nu = 1/m$ the exact $q^4$ coefficient is
$S_4 = (m-1)/8$, which equals $|\mathcal{S}-1|/8$ with $\mathcal{S} = m$. The
model wavefunction saturates the bound exactly. **[Model]** — this is a
property of the trial wavefunction, not of the Coulomb Hamiltonian.

```{r laughlin_saturation}
laughlin <- data.frame(m = c(1, 3, 5, 7)) %>%
  mutate(nu       = 1 / m,
         S        = m,
         S4_exact = (m - 1) / 8,
         S4_bound = abs(S - 1) / 8,
         saturated = isTRUE(all.equal(S4_exact, S4_bound)))

knitr::kable(laughlin, digits = 4,
             caption = "Laughlin states saturate Haldane's bound exactly.")

stopifnot(all(abs(laughlin$S4_exact - laughlin$S4_bound) < 1e-12))
```

<br />

# The chiral spectral sum rules

Define two non-negative spectral densities $\rho_T^{(-)}(\omega)$ and
$\rho_T^{(+)}(\omega)$, proportional to the absorption probabilities for
gravitons of chirality $-2$ and $+2$ respectively. Their inverse-frequency
moments,

$$\mathcal{J}_\mp \;=\; \int_0^\infty \frac{d\omega}{2\pi}\,
\frac{\rho_T^{(\mp)}(\omega)}{\omega},$$

satisfy two sum rules. **[Convention]** — written here in the normalization
that makes Haldane's bound come out as stated; verify factors against
Golkar–Nguyen–Son before quoting numerically:

$$\mathcal{J}_- - \mathcal{J}_+ \;=\; \frac{\mathcal{S} - 1}{8},
\qquad\qquad
\mathcal{J}_- + \mathcal{J}_+ \;=\; S_4 .$$

The first is **topological** — fixed by the shift, independent of the
Hamiltonian. The second is **dynamical** — it depends on the actual ground
state. Because $\mathcal{J}_\pm \ge 0$ by construction, the two together give
Haldane's bound and, crucially, tell us that *saturation is equivalent to one
channel being empty*.

Define the **chirality purity**

$$P \;=\; \frac{\mathcal{J}_- - \mathcal{J}_+}{\mathcal{J}_- + \mathcal{J}_+}
\;=\; \frac{\mathcal{S} - 1}{8\,S_4}, \qquad P \in [-1, 1].$$

$|P| = 1$ is maximal chirality. This single number is what the CP-RILS
polarization asymmetry is actually measuring.

```{r sum_rules}
chiral_weights <- function(S, S4) {
  d  <- (S - 1) / 8
  Jm <- (S4 + d) / 2
  Jp <- (S4 - d) / 2
  if (Jm < -1e-12 || Jp < -1e-12)
    warning("S4 violates Haldane's bound for this shift; no valid state.")
  data.frame(S = S, S4 = S4, J_minus = Jm, J_plus = Jp,
             purity = (Jm - Jp) / (Jm + Jp))
}

scenarios <- bind_rows(
  cbind(case = "nu=1/3, Laughlin model wf",   chiral_weights(3, 1/4)),
  cbind(case = "nu=1/3, Coulomb (S4 larger)", chiral_weights(3, 0.30)),
  cbind(case = "nu=2/3, anti-Laughlin",       chiral_weights(0, 1/8)),
  cbind(case = "nu=2/7, single graviton",     chiral_weights(2, 1/8)),
  cbind(case = "nu=2/7, plus parton mode",    chiral_weights(2, 1/2))
)

knitr::kable(scenarios, digits = 4,
             caption = "Chiral spectral weights and purity. Purity = +/-1 means one polarization channel is empty.")
```

## Reading the table

- **$\nu = 1/3$, model wavefunction**: $P = +1$. All weight in the $-2$
  channel. This is the idealized statement.
- **$\nu = 1/3$, Coulomb**: numerics on Coulomb ground states find $S_4$
  *above* the bound even at $\nu = 1/3$ and $1/5$, so $|P| < 1$ and both
  chiralities carry weight. **[Established, numerically]** The 2024 experiment
  therefore demonstrates *near*-maximal chirality, not exactly maximal. The
  source transcript does not make this distinction.
- **$\nu = 2/7$**: $\mathcal{S} = 2$, so the topological sum rule fixes the
  *net* weight at only $1/8$ — a small number. If the measured $S_4$ is
  substantially larger, the excess is forced into the *opposite*-chirality
  channel. That is precisely the room the second, high-energy graviton
  occupies. The 2026 multi-graviton observation is not merely compatible
  with the sum rules; it is the kind of thing the sum rules make room for.
  This is the Nguyen–Son argument: a Dirac composite-fermion description of
  $\nu = n/(4n \pm 1)$ violates Haldane's bound unless an additional mode is
  added. **[Standard theory]**

```{r purity_figure, fig.height=5.5}
grid <- expand.grid(S  = c(0, 2, 3, 4, 6),
                    S4 = seq(0.01, 1.0, by = 0.005)) %>%
  rowwise() %>%
  mutate(purity = (S - 1) / (8 * S4)) %>%
  ungroup() %>%
  filter(abs(purity) <= 1) %>%
  mutate(state = factor(S, levels = c(0, 2, 3, 4, 6),
                        labels = c("nu=2/3 (S=0)", "nu=2/7 (S=2)",
                                   "nu=1/3 (S=3)", "nu=2/5 (S=4)",
                                   "nu=2/9 (S=6)")))

ggplot(grid, aes(x = S4, y = purity, colour = state)) +
  geom_line(aes(colour = state), size = 1.1) +
  geom_hline(yintercept = c(-1, 0, 1), linetype = "dashed",
             colour = "grey45") +
  annotate("text", x = 0.92, y = 0.93, label = "maximally chiral (-2)",
           size = 3.2, colour = "grey25") +
  annotate("text", x = 0.92, y = -0.93, label = "maximally chiral (+2)",
           size = 3.2, colour = "grey25") +
  scale_colour_viridis_d(name = "FQH state", end = 0.9) +
  labs(
    title = "Graviton chirality purity is fixed by the shift and S4",
    subtitle = "P = (S - 1) / (8 * S4); each curve terminates where Haldane's bound is saturated",
    x = "S4  (q^4 coefficient of the projected static structure factor)",
    y = "Chirality purity P"
  ) +
  theme(legend.position = "right")
```

The left endpoint of each curve is the Haldane bound for that state. Moving
right along a curve — increasing $S_4$ at fixed topology, i.e. deforming away
from the model wavefunction toward realistic Coulomb interactions — pushes
$|P|$ down and populates the opposite polarization channel. Note that
$\nu = 2/7$ ($\mathcal{S}=2$) loses purity fastest, which is why it is the
natural place to look for a second graviton.

<br />

# Long-wavelength structure factor and the $\nu = 1$ control

The projected static structure factor $\bar{S}(q)$ starts at fourth order:

$$\bar{S}(q\ell_B) \;=\; S_4 (q\ell_B)^4 \;+\; S_6 (q\ell_B)^6 \;+\; \dots$$

**Why this matters experimentally.** CP-RILS transfers momentum
$q\ell_B \ll 1$. The experiment lives entirely inside the regime where the
exact $S_4$ coefficient dominates — it is not probing the roton minimum near
$q\ell_B \approx 1.4$. This is the direct link between the measurement and the
topological bookkeeping above.

**A caution on $S_6$.** **[Corrected / Conjectural]** Effective-field-theory
predictions for $S_6$ have been checked numerically and do *not* hold as
equalities for generic FQH states. Do not treat $S_6$ as universal.

```{r structure_factor}
x <- seq(0, 1.2, length.out = 300)   # x = q * l_B

sf_df <- fqh_states %>%
  filter(label %in% c("1", "1/3", "2/5", "1/5")) %>%
  select(label, S4_bound) %>%
  tidyr::crossing(x = x) %>%
  mutate(Sbar = S4_bound * x^4)

ggplot(sf_df, aes(x = x, y = Sbar, colour = label)) +
  geom_line(aes(colour = label), size = 1.1) +
  scale_colour_viridis_d(name = "Filling", end = 0.85) +
  labs(
    title = "Leading long-wavelength projected structure factor",
    subtitle = "Model-wavefunction values (bound saturated). The nu = 1 curve is identically zero.",
    x = expression(q * l[B]),
    y = expression(bar(S)(q))
  )
```

**Exact control calculation.** For $\nu = 1$ the full static structure factor
is $S(q) = 1 - e^{-q^2\ell_B^2/2}$, and the projection subtracts exactly that
term, giving $\bar{S}(q) \equiv 0$. No geometric degree of freedom, no
graviton. This is an analytic benchmark, not a fit. **[Established]**

```{r iqhe_control}
S_full_nu1 <- function(x) 1 - exp(-x^2 / 2)
S_proj_nu1 <- function(x) S_full_nu1(x) - (1 - exp(-x^2 / 2))

max_dev <- max(abs(S_proj_nu1(seq(0, 5, length.out = 1001))))
cat(sprintf("Max |Sbar(q)| for nu = 1 over q*l_B in [0,5]:  %.3e\n", max_dev))
stopifnot(max_dev < 1e-12)
```

The single-mode approximation dispersion is
$\Delta(q) = \bar{f}(q) / \bar{S}(q)$, with $\bar{f}(q)$ the projected
oscillator strength. Evaluating it requires the Laughlin pair correlation
function $g(r)$, which is **not** computed here; a hypernetted-chain or
sphere-geometry exact-diagonalization implementation is deferred (see
Roadmap).

<br />

# Polarization selection rules

The measurement is angular-momentum bookkeeping on photons. Writing $L_{\rm
in}$ and $L_{\rm out}$ for the photon angular momenta along the field axis,
the transfer to the electron liquid is $\Delta L = L_{\rm in} - L_{\rm out}$.

```{r selection_rules}
sel <- expand.grid(L_in = c(-1, 1), L_out = c(-1, 1)) %>%
  mutate(
    dL      = L_in - L_out,
    channel = ifelse(dL == 0, "spin-0 (density)", "spin-2 (graviton)"),
    couples = case_when(
      dL ==  0 ~ "charge-density excitations",
      dL == -2 ~ "chiral graviton, chirality -2",
      dL == +2 ~ "chiral graviton, chirality +2"
    )
  )

knitr::kable(sel, digits = 0,
             col.names = c("L_in", "L_out", "Delta L", "Channel", "Couples to"),
             caption = "CP-RILS angular momentum bookkeeping.")
```

At $\nu = 1/3$ the signal appears only in the $\Delta L = -2$ channel.
**[Established]**

**[Convention]** Mapping optical *handedness* ($\sigma^\pm$) onto photon
angular momentum $L = \pm 1$ depends on propagation direction. In a
backscattering geometry the outgoing photon travels along $-z$, which flips
the handedness-to-$L_z$ assignment. State results in terms of $L$, not
handedness, or the sign of the reported chirality can invert. This is a real
source of confusion in the secondary literature.

<br />

# What the CGM is not

The Weinberg-type argument reproduced in `Claude_AI_Science_Chats.txt`
("Spin-2 Particle and the Mathematics of Gravity") concludes that a consistent
theory of *massless* spin-2 particles in four dimensions coupling to a
conserved $T^{\mu\nu}$ must be general relativity. It is worth being explicit
that **none of the premises hold** for the CGM. **[Established]**

| Property of a fundamental graviton | CGM in an FQH liquid |
|---|---|
| Massless | **Gapped** — the mode sits at the magnetoroton gap |
| Both helicities $\pm 2$ | **Single dominant chirality**, set by the shift |
| Spin-2 of the Lorentz group $SO(3,1)$ | Charge-2 of the planar rotation group $SO(2)$ |
| Universal coupling to $T^{\mu\nu}$ | Couples to the internal guiding-centre metric only |
| Diffeomorphism invariance | Absent; the host lattice and $\mathbf{B}$ pick a frame |
| Long-range $1/r^2$ force | None; the mode is gapped and neutral |
| Lorentz invariance | Absent; non-relativistic, strong magnetic field |
| Self-interacting per Einstein–Hilbert | No analogue established |

So the shared structure is exactly one quantum number. The physics that *is*
genuinely transferable is the **response to a background metric
perturbation** — Hall viscosity, $\eta_H = \tfrac{\hbar}{4}\bar{s}\,\bar{n}$
(Read–Rezayi), is a bona fide geometric response function and is the honest
bridge between this material and gravitational physics. **[Standard theory]**

<br />

# Connections to the repository

## Genuine technical dependencies

**`Gravitons_Gluons_Branes.md` — measuring gravitational quantities with
non-gravitational probes.** The proton's gravitational form factors
$A(t), B(t), D(t)$ are matrix elements of $T^{\mu\nu}$ extracted from DVCS and
near-threshold $J/\psi$ photoproduction — electromagnetic experiments
delivering the proton's coupling to a graviton. The CGM measurement is the
same methodological move in condensed matter: a spin-2 stress-tensor
observable extracted from an optical probe. In both cases the operator, not
the force, is what is accessed. This is a real parallel, not decorative.

**Hall viscosity and the $D$-term.** Both are stress-response coefficients
obtained from the same energy–momentum operator; $\bar{s}$ plays the role in
FQH that the orbital-spin decomposition plays in Ji's sum rule. Worth a
side-by-side treatment.

**`Fracton_Codes_in_R.Rmd` — the guiding-centre algebra.** The LLL projected
density operators close on the GMP algebra, a $W_\infty$ / area-preserving
diffeomorphism algebra. Dipole-conserving field theories — the parent
structures of fracton mobility restrictions — arise from the same
multipole-constraint logic (Gromov; Nguyen–Gromov–Moroz; Prem–Pretko). The
mobility restriction of a fracton and the LLL constraint that ties an
excitation's dipole moment to its position are the *same* algebraic
mechanism. **Genuine dependency.**

Pretko's "emergent gravity from symmetric tensor gauge theory" is a
*different* and much weaker claim, and should not be merged with the above.
**Structural analogy, flag as such.**

## Structural analogies only

- **`Quantum_Chromodynamics_in_R.Rmd` — partons.** False cognate, as noted
  above. Flag explicitly wherever both documents are cited together.
- **`Gravitation.md` — Scherk's graviphoton.** Both concern spin content of
  gravitational multiplets, but the graviphoton is a spin-1 field in extended
  supergravity, and no FQH construction produces it. No dependency.
- **`Quasiparticle_Rectification.md` — energy transport.** The CGM is neutral
  and gapped, so it carries energy without charge. Whether that is useful for
  anything is unestablished; the thermodynamic constraints in that document
  are unaffected. **[Conjectural]** — do not develop without a concrete
  transport calculation.

## Frontier items for `QuasiParticles_in_R.Rmd`

Chiral graviton modes have moved into fractional Chern insulators and moiré
systems during 2026:

- Long, Lu, Wu, Meng, *Phys. Rev. B* **113**, L041108 (2026) — magnetoroton
  and CGM spectra in fractional Chern insulators.
- Shen et al., *Phys. Rev. B* **113**, L081403 (2026) — magnetorotons in
  moiré fractional Chern insulators.
- CGMs in non-Abelian lattice FQH states (bosonic Harper–Hofstadter at unit
  filling, Moore–Read ground state), arXiv:2607.06267.

The Moore–Read / anti-Pfaffian discrimination by graviton chirality is the
sharpest open experimental target: the two candidates at $\nu = 5/2$ carry
*opposite* graviton chirality, so CP-RILS is in principle a direct
topological-order discriminator. **[Standard theory, experimentally open]**

<br />

# Roadmap

1. **Sphere-geometry exact diagonalization** of the hard-core ($V_1$)
   pseudopotential Hamiltonian at $\nu = 1/3$ for $N = 4$–$6$ electrons.
   $N_\phi = 3(N-1)$, so $N=6$ gives $\binom{16}{6} = 8008$ basis states —
   comfortably in range for R. Extract the exact Laughlin ground state, the
   $L = 2$ graviton, and $\bar{S}(q)$ directly. This would make $S_4$ a
   *computed* rather than *quoted* quantity and would verify the saturation
   claim numerically. Requires Wigner $3j$ symbols and Haldane
   pseudopotentials.

2. **Cross-check the Coulomb non-saturation result** (arXiv:2304.14991) at
   $\nu = 1/3$ by comparing $V_1$-only against a truncated Coulomb
   pseudopotential set — the same "compare against a simpler control model"
   pattern used in `QET_Subsystem_Symmetry_in_R.Rmd`.

3. **GMP single-mode approximation dispersion** once $g(r)$ is available from
   step 1; recover the roton minimum near $q\ell_B \approx 1.4$.

4. **Verify the sum-rule normalization** in §"The chiral spectral sum rules"
   against Golkar–Nguyen–Son eq. numbering, and remove the **[Convention]**
   flag once confirmed.

5. **Hall viscosity / $D$-term comparison section** bridging this document
   and `Gravitons_Gluons_Branes.md`.

<br />

# References

**Experiment**

- Liang, J., Liu, Z., Yang, Z., Huang, Y., Wurstbauer, U., Dean, C. R.,
  West, K. W., Pfeiffer, L. N., Du, L., & Pinczuk, A. (2024). Evidence for
  chiral graviton modes in fractional quantum Hall liquids. *Nature*
  **628**, 78–83.
- Yang, Z., Wang, Y., Lu, X., Liu, Z., Baldwin, K. W., West, K. W.,
  Pfeiffer, L. N., Yang, B., Balram, A. C., & Du, L. (2026). Emergent
  partons in fractional quantum Hall systems. *Nature Physics*,
  doi:10.1038/s41567-026-03338-9.
- Pinczuk, A., Dennis, B. S., Pfeiffer, L. N., & West, K. W. (1993).
  Observation of collective excitations in the fractional quantum Hall
  effect. *Phys. Rev. Lett.* **70**, 3983.

**Theory**

- Girvin, S. M., MacDonald, A. H., & Platzman, P. M. (1985). Collective-
  excitation gap in the fractional quantum Hall effect. *Phys. Rev. Lett.*
  **54**, 581.
- Haldane, F. D. M. (2011). Geometrical description of the fractional
  quantum Hall effect. *Phys. Rev. Lett.* **107**, 116801.
- Golkar, S., Nguyen, D. X., & Son, D. T. (2016). Spectral sum rules and
  magneto-roton as emergent graviton in the fractional quantum Hall effect.
  *JHEP* **2016**, 21.
- Liou, S.-F., Haldane, F. D. M., Yang, K., & Rezayi, E. H. (2019). Chiral
  gravitons in fractional quantum Hall liquids. *Phys. Rev. Lett.* **123**,
  146801.
- Nguyen, D. X., & Son, D. T. (2021). Probing the spin structure of the
  fractional quantum Hall magnetoroton with polarized Raman scattering.
  *Phys. Rev. Research* **3**, 023040.
- Nguyen, D. X., Haldane, F. D. M., Rezayi, E. H., Son, D. T., & Yang, K.
  (2022). Multiple magnetorotons and spectral sum rules in fractional
  quantum Hall systems. *Phys. Rev. Lett.* **128**, 246402.
- Read, N., & Rezayi, E. H. (2011). Hall viscosity, orbital spin, and
  geometry. *Phys. Rev. B* **84**, 085316.
- Balram, A. C., Pu, S., et al. — parton excitons and very-high-energy
  collective states, *Phys. Rev. X* **12**, 021008 (2022).

**Companion documents in this repository**

`Gravitation.md` · `Gravitons_Gluons_Branes.md` ·
`Quantum_Chromodynamics_in_R.Rmd` · `Fracton_Codes_in_R.Rmd` ·
`QuasiParticles_in_R.Rmd` · `QET_Subsystem_Symmetry_in_R.Rmd`

---

*AI-assistance disclosure: the corrections, sum-rule bookkeeping, and
repository cross-references in this document were developed in an
AI-assisted session. Every numerical claim is either reproduced by the code
blocks above or flagged for verification against the cited primary source.
Items still requiring verification are listed in the Roadmap.*
