---
title: "Fracton Codes Foundations and Analysis in R"
author: "John Akwei, Senior Data Scientist"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    df_print: kable
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = 'center', cache = TRUE)
```

```{r libraries, message=FALSE}
# Load required libraries
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(gridExtra)
library(viridis)
library(reshape2)

# Set theme for plots
theme_set(theme_minimal())
```

<br />
<br />

# Abstract

Fracton codes represent a revolutionary class of quantum error-correcting codes characterized by quasiparticle excitations with severely constrained mobility. Unlike conventional topological codes where anyons can move freely, fracton codes exhibit point-like excitations that are immobile or have restricted dimensional mobility. This document provides a comprehensive mathematical analysis of fracton codes, demonstrating their unique properties through stabilizer formalism, higher-moment conservation laws, and computational implementations in R.

<br />
<br />

# Introduction: From Anyons to Fractons

## Conventional Topological Codes

In 2D topological codes like the toric code, anyonic excitations are created and manipulated by string-like Pauli operators:

$$W_\gamma = \prod_{i \in \gamma} \sigma_i^\alpha$$

These anyons can move freely in the plane, with topological charge encoded in their braiding statistics characterized by modular S and T matrices.

## The Fracton Revolution

Fracton codes break this paradigm. In 3D fracton models, isolated point-like excitations (fractons) cannot be translated by any finite-support operator. There exists no local operator $O$ such that:

$$O|vac\rangle = |vac \text{ with one fracton translated}\rangle$$

Moving a fracton necessarily creates additional excitations, leading to fundamentally constrained dynamics.

```{r fracton_vs_anyon}
# Visualize mobility constraints
mobility_data <- data.frame(
  code_type = c("2D Toric Code", "3D X-Cube", "Haah's Cubic Code", "Surface Code"),
  dimensionality = c(2, 3, 3, 2),
  excitation_mobility = c("Free 2D", "Restricted (lineons/planons)", "Immobile fractons", "Free 2D"),
  gsd_scaling = c("Constant", "Linear (L)", "Linear (L)", "Constant")
)

print("Comparison of Topological and Fracton Codes:")
print(mobility_data)

ggplot(mobility_data, aes(x = reorder(code_type, -dimensionality), 
                          y = dimensionality, fill = excitation_mobility)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(title = "Code Dimensionality and Excitation Mobility",
       x = "Code Type",
       y = "Spatial Dimensionality",
       fill = "Excitation\nMobility") +
  scale_fill_viridis_d()
```

## Ground State Degeneracy Scaling

A key signature of fracton order is the subextensive scaling of ground state degeneracy:

$$\log D(L) \sim cL$$

for a system of linear size $L$, contrasting sharply with the constant degeneracy of conventional 2D topological order.

```{r gsd_scaling}
# Simulate ground state degeneracy scaling
L_values <- seq(10, 100, by = 5)

# Different scaling behaviors
toric_gsd <- rep(4, length(L_values))  # Constant (2^2 for toric code)
xcube_gsd <- 2 * L_values  # Linear scaling
haah_gsd <- 1.5 * L_values  # Linear with different coefficient

gsd_data <- data.frame(
  L = rep(L_values, 3),
  log_D = c(log2(toric_gsd), log2(xcube_gsd), log2(haah_gsd)),
  code = rep(c("Toric Code (2D)", "X-Cube Model", "Haah's Cubic Code"), 
             each = length(L_values))
)

ggplot(gsd_data, aes(x = L, y = log_D, color = code)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Ground State Degeneracy Scaling",
       subtitle = "Fracton codes exhibit subextensive scaling",
       x = "System Linear Size (L)",
       y = "log₂(D) - Number of Logical Qubits",
       color = "Code Type") +
  scale_color_manual(values = c("blue", "red", "green")) +
  annotate("text", x = 70, y = 2.5, 
           label = "Topological:\nConstant", color = "blue", size = 3.5) +
  annotate("text", x = 70, y = 100, 
           label = "Fracton:\nLinear ~ L", color = "red", size = 3.5)
```

<br />
<br />

# Mathematical Framework: Stabilizer Formalism

## General Stabilizer Structure

Fracton codes are exactly solvable spin Hamiltonians in stabilizer form:

$$H = -\sum_a J_a S_a, \quad S_a \in \mathcal{P}^{\otimes N}, \quad S_a^2 = 1, \quad [S_a, S_b] = 0$$

The code space is the common +1 eigenspace of all stabilizers $S_a$.

## CSS Code Structure

Many fracton codes are CSS (Calderbank-Shor-Steane) codes with separate X-type and Z-type stabilizers:

$$H = -\sum_v A_v - \sum_c B_c$$
$$A_v = \prod_{i \in v} X_i, \quad B_c = \prod_{i \in c} Z_i$$

```{r stabilizer_framework}
# Implement basic stabilizer operations
create_pauli_operator <- function(qubits, operator_type, support) {
  # Create a Pauli operator on specified qubits
  # operator_type: "X", "Y", or "Z"
  # support: vector of qubit indices
  
  n_qubits <- qubits
  operator <- rep("I", n_qubits)
  operator[support] <- operator_type
  
  return(list(
    operator = operator,
    support = support,
    weight = length(support)
  ))
}

# Example: Create stabilizers for a small fracton code
n_qubits <- 8
stabilizers <- list()

# X-type stabilizers (acting on vertices)
stabilizers[[1]] <- create_pauli_operator(n_qubits, "X", c(1, 2, 3, 4))
stabilizers[[2]] <- create_pauli_operator(n_qubits, "X", c(5, 6, 7, 8))

# Z-type stabilizers (acting on faces)
stabilizers[[3]] <- create_pauli_operator(n_qubits, "Z", c(1, 2, 5, 6))
stabilizers[[4]] <- create_pauli_operator(n_qubits, "Z", c(3, 4, 7, 8))

# Display stabilizer structure
stabilizer_df <- data.frame(
  index = 1:length(stabilizers),
  type = c("X", "X", "Z", "Z"),
  support = sapply(stabilizers, function(s) paste(s$support, collapse = ",")),
  weight = sapply(stabilizers, function(s) s$weight)
)

print("Stabilizer Generator Structure:")
print(stabilizer_df)

# Visualize stabilizer weights
ggplot(stabilizer_df, aes(x = index, y = weight, fill = type)) +
  geom_col(alpha = 0.8) +
  labs(title = "Stabilizer Generator Weights",
       x = "Stabilizer Index",
       y = "Weight (Number of Qubits)",
       fill = "Pauli Type") +
  scale_fill_manual(values = c("X" = "red", "Z" = "blue"))
```

## Excitations and Defects

Excitations correspond to stabilizer violations:

- An eigenvalue $S_a = -1$ marks a defect
- Defects are interpreted as fractons or subdimensional excitations
- Local errors flip stabilizers, creating excitation clusters

```{r excitation_dynamics}
# Simulate excitation creation and dynamics
create_excitation_cluster <- function(lattice_size, error_location) {
  # Create a pattern of violated stabilizers from a local error
  lattice <- array(0, dim = c(lattice_size, lattice_size, lattice_size))
  
  # Mark violated stabilizers near error
  x <- error_location[1]
  y <- error_location[2]
  z <- error_location[3]
  
  # Simple model: error affects nearby stabilizers
  for (dx in -1:1) {
    for (dy in -1:1) {
      for (dz in -1:1) {
        nx <- x + dx
        ny <- y + dy
        nz <- z + dz
        if (nx >= 1 && nx <= lattice_size &&
            ny >= 1 && ny <= lattice_size &&
            nz >= 1 && nz <= lattice_size) {
          lattice[nx, ny, nz] <- 1
        }
      }
    }
  }
  
  return(lattice)
}

# Generate excitation pattern
L <- 8
error_pos <- c(4, 4, 4)
excitations <- create_excitation_cluster(L, error_pos)

# Count total excitations
n_excitations <- sum(excitations)

cat("Single error creates", n_excitations, "violated stabilizers\n")
cat("Energy cost: E =", 2 * n_excitations, "J (assuming uniform coupling)\n")

# Visualize 2D slice of excitation pattern
slice_z <- 4
excitation_slice <- excitations[, , slice_z]
excitation_df <- expand.grid(x = 1:L, y = 1:L)
excitation_df$violated <- as.vector(excitation_slice)

ggplot(excitation_df, aes(x = x, y = y, fill = factor(violated))) +
  geom_tile(color = "white") +
  labs(title = paste("Excitation Pattern (z =", slice_z, "slice)"),
       x = "x coordinate", y = "y coordinate",
       fill = "Stabilizer\nViolated") +
  scale_fill_manual(values = c("0" = "white", "1" = "red")) +
  coord_equal()
```

<br />
<br />

# Logical Operators: Membranes and Fractals

## Structure of Logical Operators

In fracton codes, logical operators take membrane or fractal forms:

$$\bar{X} = \prod_{i \in M} X_i, \quad \bar{Z} = \prod_{i \in F} Z_i$$

where $M$ is a 2D membrane and $F$ can be a fractal subset of the lattice.

```{r logical_operators}
# Implement membrane and fractal logical operators
create_membrane_operator <- function(lattice_size, plane = "xy", position = NULL) {
  # Create a membrane operator (2D surface)
  membrane <- array(0, dim = c(lattice_size, lattice_size, lattice_size))
  
  if (plane == "xy") {
    z_pos <- ifelse(is.null(position), lattice_size/2, position)
    membrane[, , z_pos] <- 1
  } else if (plane == "xz") {
    y_pos <- ifelse(is.null(position), lattice_size/2, position)
    membrane[, y_pos, ] <- 1
  } else if (plane == "yz") {
    x_pos <- ifelse(is.null(position), lattice_size/2, position)
    membrane[x_pos, , ] <- 1
  }
  
  return(membrane)
}

create_fractal_operator <- function(lattice_size, depth = 3) {
  # Create a simplified fractal pattern (Sierpinski-like)
  fractal <- array(0, dim = c(lattice_size, lattice_size, lattice_size))
  
  # Recursive subdivision pattern
  for (i in 1:lattice_size) {
    for (j in 1:lattice_size) {
      for (k in 1:lattice_size) {
        # Simple rule: include if sum of binary digits is even
        if ((sum(as.integer(intToBits(i)[1:3])) + 
             sum(as.integer(intToBits(j)[1:3])) + 
             sum(as.integer(intToBits(k)[1:3]))) %% 2 == 0) {
          fractal[i, j, k] <- 1
        }
      }
    }
  }
  
  return(fractal)
}

# Generate examples
L <- 16
membrane_xy <- create_membrane_operator(L, "xy", L/2)
fractal_pattern <- create_fractal_operator(L, depth = 3)

# Calculate weights
membrane_weight <- sum(membrane_xy)
fractal_weight <- sum(fractal_pattern)

cat("Membrane operator weight:", membrane_weight, "\n")
cat("Fractal operator weight:", fractal_weight, "\n")
cat("Membrane dimension: 2D (L²)\n")
cat("Fractal dimension: ~", log(fractal_weight)/log(L), "\n")

# Visualize membrane operator
membrane_slice <- membrane_xy[, , L/2]
membrane_df <- expand.grid(x = 1:L, y = 1:L)
membrane_df$operator <- as.vector(membrane_slice)

p1 <- ggplot(membrane_df, aes(x = x, y = y, fill = factor(operator))) +
  geom_tile(color = "gray90") +
  labs(title = "Membrane Logical Operator",
       subtitle = "2D surface in 3D lattice") +
  scale_fill_manual(values = c("0" = "white", "1" = "darkblue")) +
  coord_equal() +
  theme(legend.position = "none")

# Visualize fractal operator
fractal_slice <- fractal_pattern[, , L/2]
fractal_df <- expand.grid(x = 1:L, y = 1:L)
fractal_df$operator <- as.vector(fractal_slice)

p2 <- ggplot(fractal_df, aes(x = x, y = y, fill = factor(operator))) +
  geom_tile(color = "gray90") +
  labs(title = "Fractal Logical Operator",
       subtitle = "Self-similar pattern") +
  scale_fill_manual(values = c("0" = "white", "1" = "darkred")) +
  coord_equal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

<br />
<br />

# Higher-Moment Conservation Laws

## Rank-2 U(1) Gauge Theory

At the field-theoretic level, fracton phases can be modeled by rank-2 gauge theories with Gauss-law constraints:

$$\partial_i \partial_j E^{ij} = \rho$$

where $\rho$ is fracton charge density and $E^{ij}$ is a symmetric tensor electric field.

## Multipole Conservation

Integrating the constraint yields:

**Total charge conservation:**
$$Q = \int d^d x \, \rho(x)$$

**Dipole moment conservation:**
$$P_k = \int d^d x \, x_k \rho(x)$$

```{r multipole_conservation}
# Demonstrate multipole conservation constraints
calculate_multipoles <- function(charges, positions) {
  # charges: vector of charge values
  # positions: matrix of positions (n_charges x 3)
  
  Q <- sum(charges)  # Total charge
  
  # Dipole moment
  P <- colSums(charges * positions)
  
  # Quadrupole moment tensor (simplified)
  Q_tensor <- matrix(0, nrow = 3, ncol = 3)
  for (i in 1:nrow(positions)) {
    r <- positions[i, ]
    Q_tensor <- Q_tensor + charges[i] * outer(r, r)
  }
  
  return(list(
    total_charge = Q,
    dipole = P,
    quadrupole = Q_tensor
  ))
}

# Example: Moving a single charge violates dipole conservation
initial_pos <- matrix(c(0, 0, 0), nrow = 1)
initial_charge <- c(1)

final_pos <- matrix(c(5, 0, 0), nrow = 1)
final_charge <- c(1)

multipoles_initial <- calculate_multipoles(initial_charge, initial_pos)
multipoles_final <- calculate_multipoles(final_charge, final_pos)

cat("Initial state:\n")
cat("  Q =", multipoles_initial$total_charge, "\n")
cat("  P = (", paste(multipoles_initial$dipole, collapse = ", "), ")\n\n")

cat("After moving single fracton:\n")
cat("  Q =", multipoles_final$total_charge, "\n")
cat("  P = (", paste(multipoles_final$dipole, collapse = ", "), ")\n\n")

delta_P <- multipoles_final$dipole - multipoles_initial$dipole
cat("Change in dipole: ΔP = (", paste(delta_P, collapse = ", "), ")\n")
cat("This violates dipole conservation!\n\n")

# Now show a dipole configuration that can move
dipole_pos <- matrix(c(0, 0, 0,
                       1, 0, 0), nrow = 2, byrow = TRUE)
dipole_charges <- c(1, -1)

dipole_pos_moved <- matrix(c(5, 0, 0,
                              6, 0, 0), nrow = 2, byrow = TRUE)

multipoles_dipole_init <- calculate_multipoles(dipole_charges, dipole_pos)
multipoles_dipole_final <- calculate_multipoles(dipole_charges, dipole_pos_moved)

cat("Dipole configuration (charge neutral):\n")
cat("  Initial P = (", paste(multipoles_dipole_init$dipole, collapse = ", "), ")\n")
cat("  Final P = (", paste(multipoles_dipole_final$dipole, collapse = ", "), ")\n")
cat("  Dipole moment CONSERVED - motion allowed!\n")
```

## Mobility Constraints Visualization

If total charge $Q$ and dipole $P_k$ are both conserved, translating an isolated charge by vector $\mathbf{a}$ would change:

$$\Delta P_k = Q a_k$$

This is forbidden unless additional charges are created.

```{r mobility_visualization}
# Visualize allowed vs forbidden moves
moves_data <- data.frame(
  configuration = c("Single fracton", "Dipole (neutral)", "Quadrupole", 
                    "Line of fractons", "Isolated fracton pair"),
  charge_Q = c(1, 0, 0, 4, 2),
  dipole_conserved = c("No", "Yes", "Yes", "Partially", "No"),
  mobility = c("Immobile", "1D mobile", "Limited", "1D mobile (lineon)", "Immobile"),
  allowed = c(FALSE, TRUE, TRUE, TRUE, FALSE)
)

print("Mobility Classification by Multipole Configuration:")
print(moves_data)

ggplot(moves_data, aes(x = reorder(configuration, -charge_Q), 
                       y = charge_Q, fill = allowed)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(title = "Charge Configurations and Mobility Constraints",
       x = "Configuration Type",
       y = "Total Charge Q",
       fill = "Motion\nAllowed") +
  scale_fill_manual(values = c("FALSE" = "red", "TRUE" = "green")) +
  geom_text(aes(label = mobility), hjust = -0.1, size = 3)
```

<br />
<br />

# Type I Fracton Codes: The X-Cube Model

## Hamiltonian Structure

The X-cube model Hamiltonian is:

$$H_{X\text{-cube}} = -\sum_c A_c - \sum_{v,\mu} B_v^\mu$$

where:
- $A_c$ are cube stabilizers (X operators on edges of cube $c$)
- $B_v^\mu$ are cross-shaped Z-type stabilizers in planes through vertex $v$ normal to direction $\mu \in \{x, y, z\}$

```{r xcube_model}
# Implement X-cube model structure
create_xcube_lattice <- function(Lx, Ly, Lz) {
  # Create 3D cubic lattice with qubits on edges
  # Each edge has one qubit
  
  n_edges <- Lx * Ly * (Lz + 1) +  # z-direction edges
             Lx * (Ly + 1) * Lz +  # y-direction edges
             (Lx + 1) * Ly * Lz    # x-direction edges
  
  return(list(
    Lx = Lx, Ly = Ly, Lz = Lz,
    n_edges = n_edges,
    n_vertices = (Lx + 1) * (Ly + 1) * (Lz + 1),
    n_cubes = Lx * Ly * Lz
  ))
}

# Generate X-cube lattice
L <- 4
xcube <- create_xcube_lattice(L, L, L)

cat("X-Cube Model Parameters:\n")
cat("  Lattice size:", L, "×", L, "×", L, "\n")
cat("  Number of qubits (edges):", xcube$n_edges, "\n")
cat("  Number of vertices:", xcube$n_vertices, "\n")
cat("  Number of cubes:", xcube$n_cubes, "\n")
cat("  Number of stabilizers:\n")
cat("    Cube stabilizers (A_c):", xcube$n_cubes, "\n")
cat("    Vertex stabilizers (B_v^μ):", 3 * xcube$n_vertices, "\n")

# Calculate code parameters
n_stabilizers <- xcube$n_cubes + 3 * xcube$n_vertices
k_logical <- xcube$n_edges - n_stabilizers  # Logical qubits

cat("  Logical qubits:", k_logical, "\n")
cat("  Ground state degeneracy: 2^", k_logical, " = ", 2^k_logical, "\n")

# Analyze scaling
L_range <- 2:10
scaling_data <- data.frame(L = L_range)

for (i in 1:length(L_range)) {
  Ltmp <- L_range[i]
  xcube_tmp <- create_xcube_lattice(Ltmp, Ltmp, Ltmp)
  n_stab <- xcube_tmp$n_cubes + 3 * xcube_tmp$n_vertices
  k_log <- xcube_tmp$n_edges - n_stab
  scaling_data$k[i] <- k_log
}

ggplot(scaling_data, aes(x = L, y = k)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "X-Cube Model: Logical Qubits vs System Size",
       subtitle = "Linear scaling: k ~ L",
       x = "Linear System Size (L)",
       y = "Number of Logical Qubits (k)") +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red")
```

## Excitation Types

The X-cube model supports multiple excitation types:

- **Fractons**: Immobile point-like excitations at cube corners
- **Lineons**: Excitations mobile along 1D lines
- **Planons**: Excitations mobile within 2D planes

```{r xcube_excitations}
# Classify excitation types by mobility
excitation_types <- data.frame(
  type = c("Fracton", "Lineon", "Planon", "Fully mobile particle"),
  mobility_dimensions = c(0, 1, 2, 3),
  example = c("Corner of violated cube", "Edge defect", "Face defect", "Not in X-cube"),
  creation_operator = c("Membrane corner", "Line operator", "Plane operator", "N/A"),
  multiplicity = c(4, "Variable", "Variable", 0)
)

print("X-Cube Model Excitation Spectrum:")
print(excitation_types)

ggplot(excitation_types[1:3, ], 
       aes(x = type, y = mobility_dimensions, fill = type)) +
  geom_col(alpha = 0.8) +
  labs(title = "Excitation Mobility in X-Cube Model",
       x = "Excitation Type",
       y = "Mobility Dimensions",
       fill = "Type") +
  scale_fill_viridis_d() +
  coord_flip() +
  theme(legend.position = "none")
```

<br />
<br />

# Type II Fracton Codes: Haah's Cubic Code

## Polynomial Construction

Haah's cubic code uses translationally invariant stabilizers defined by polynomial constraints. The generators are:

$$S_r^{(X)} = X_{r,1} X_{r,2} X_{r+\hat{x},2} X_{r+\hat{y},1} X_{r+\hat{z},2} X_{r+\hat{x}+\hat{y}+\hat{z},1}$$

$$S_r^{(Z)} = Z_{r,1} Z_{r,2} Z_{r+\hat{x},2} Z_{r+\hat{y},1} Z_{r+\hat{z},2} Z_{r+\hat{x}+\hat{y}+\hat{z},1}$$

with two qubits per site $(r, 1/2)$.

```{r haah_cubic_code}
# Implement Haah's cubic code structure
create_haah_lattice <- function(L) {
  # Each site has 2 qubits
  n_sites <- L^3
  n_qubits <- 2 * n_sites
  
  # Each site has 2 stabilizers (X and Z type)
  n_stabilizers <- 2 * n_sites
  
  return(list(
    L = L,
    n_sites = n_sites,
    n_qubits = n_qubits,
    n_stabilizers = n_stabilizers
  ))
}

# Haah code polynomial pattern
# Stabilizer acts on 6 qubits in a specific pattern
haah_pattern <- function() {
  # Relative positions in the stabilizer
  # Format: (site_offset, qubit_index)
  pattern <- data.frame(
    site = c("(0,0,0)", "(0,0,0)", "(1,0,0)", "(0,1,0)", "(0,0,1)", "(1,1,1)"),
    qubit = c(1, 2, 2, 1, 2, 1),
    position = c("origin", "origin", "+x", "+y", "+z", "+xyz")
  )
  return(pattern)
}

haah_stab_pattern <- haah_pattern()
print("Haah's Cubic Code Stabilizer Pattern:")
print(haah_stab_pattern)

# Analyze code parameters
L_values <- seq(3, 9, by = 2)
haah_params <- data.frame(L = L_values)

for (i in 1:length(L_values)) {
  Ltmp <- L_values[i]
  haah_tmp <- create_haah_lattice(Ltmp)
  
  # For Haah's code, k depends on boundary conditions
  # With periodic boundaries: k ~ L (linear)
  haah_params$n_qubits[i] <- haah_tmp$n_qubits
  haah_params$k_approx[i] <- Ltmp  # Approximate scaling
}

ggplot(haah_params, aes(x = L, y = k_approx)) +
  geom_line(color = "purple", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Haah's Cubic Code: Logical Qubits Scaling",
       subtitle = "Linear scaling k ~ αL",
       x = "Linear Lattice Size (L)",
       y = "Number of Logical Qubits (k)") +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "orange")
```

## Fractal Logical Operators

In Haah's code, separating a single fracton to infinity requires operators on fractal subsets described by polynomial zero sets over $\mathbb{F}_2$.

```{r fractal_logical}
# Demonstrate fractal structure of logical operators
generate_fractal_support <- function(L, polynomial_degree = 2) {
  # Generate points satisfying polynomial constraint over F_2
  # Simplified model of fractal support
  
  support <- expand.grid(x = 0:(L-1), y = 0:(L-1), z = 0:(L-1))
  
  # Polynomial rule (simplified): x² + y² + z² ≡ 0 (mod 2)
  support$included <- ((support$x^2 + support$y^2 + support$z^2) %% 2) == 0
  
  return(support[support$included, ])
}

# Generate fractal support
L <- 8
fractal_support <- generate_fractal_support(L)

cat("Fractal logical operator support:\n")
cat("  Total lattice sites:", L^3, "\n")
cat("  Sites in fractal support:", nrow(fractal_support), "\n")
cat("  Fractal density:", nrow(fractal_support) / L^3, "\n")

# Estimate fractal dimension
# D_f ≈ log(N) / log(L)
fractal_dim <- log(nrow(fractal_support)) / log(L)
cat("  Approximate fractal dimension:", fractal_dim, "\n")

# Visualize 2D slice of fractal
fractal_slice <- fractal_support[fractal_support$z == 4, ]

ggplot(fractal_slice, aes(x = x, y = y)) +
  geom_point(color = "darkred", size = 3, alpha = 0.8) +
  labs(title = "Fractal Logical Operator Support (z = 4 slice)",
       subtitle = paste("Haah's Cubic Code, Fractal dimension ≈", round(fractal_dim, 2)),
       x = "x coordinate", y = "y coordinate") +
  coord_equal() +
  theme_minimal() +
  xlim(0, L-1) + ylim(0, L-1)
```

<br />
<br />

# Error Dynamics and Self-Correction

## Constrained Error Propagation

The energy of an excitation cluster is:

$$E_{exc} = 2 \sum_{a \in \text{violated}} J_a$$

Local noise has difficulty redistributing excitations while preserving higher-moment constraints.

```{r error_dynamics}
# Simulate error propagation under constraints
simulate_error_spreading <- function(initial_error, timesteps, constraint_type = "fracton") {
  # Track error evolution under different constraints
  
  L <- 10
  error_positions <- list()
  error_positions[[1]] <- initial_error
  
  for (t in 2:timesteps) {
    current <- error_positions[[t-1]]
    
    if (constraint_type == "fracton") {
      # Fractons cannot move alone - errors stay localized
      error_positions[[t]] <- current
      
    } else if (constraint_type == "lineon") {
      # Lineons can diffuse along 1D
      new_pos <- current
      # Random walk along one axis
      axis <- sample(1:3, 1)
      new_pos[, axis] <- new_pos[, axis] + sample(c(-1, 1), nrow(new_pos), replace = TRUE)
      new_pos[, axis] <- pmax(1, pmin(L, new_pos[, axis]))
      error_positions[[t]] <- new_pos
      
    } else {
      # Unconstrained (2D topological code behavior)
      new_pos <- current
      for (i in 1:nrow(new_pos)) {
        step <- sample(c(-1, 0, 1), 3, replace = TRUE)
        new_pos[i, ] <- new_pos[i, ] + step
        new_pos[i, ] <- pmax(1, pmin(L, new_pos[i, ]))
      }
      error_positions[[t]] <- new_pos
    }
  }
  
  return(error_positions)
}

# Simulate different scenarios
timesteps <- 50
initial_error <- matrix(c(5, 5, 5), nrow = 1)

fracton_evolution <- simulate_error_spreading(initial_error, timesteps, "fracton")
unconstrained_evolution <- simulate_error_spreading(initial_error, timesteps, "unconstrained")

# Calculate spread over time
calculate_spread <- function(positions_list) {
  sapply(positions_list, function(pos) {
    if (nrow(pos) == 1) return(0)
    max(dist(pos))
  })
}

spread_fracton <- calculate_spread(fracton_evolution)
spread_unconstrained <- calculate_spread(unconstrained_evolution)

spread_data <- data.frame(
  time = rep(1:timesteps, 2),
  spread = c(spread_fracton, spread_unconstrained),
  type = rep(c("Fracton (immobile)", "Unconstrained (mobile)"), each = timesteps)
)

ggplot(spread_data, aes(x = time, y = spread, color = type)) +
  geom_line(size = 1.2) +
  labs(title = "Error Spreading: Fracton vs Unconstrained",
       subtitle = "Mobility constraints suppress error propagation",
       x = "Time Steps",
       y = "Error Spread (Distance)",
       color = "Code Type") +
  scale_color_manual(values = c("red", "blue"))
```

## Energy Barriers

```{r energy_barriers}
# Analyze energy barriers for error correction
calculate_energy_barrier <- function(code_type, system_size) {
  # Energy barrier scales with system size for fracton codes
  
  if (code_type == "fracton") {
    # Barrier grows with distance fracton must travel
    barrier <- system_size  # Linear in L
  } else if (code_type == "topological_2d") {
    # Barrier is constant (string tension)
    barrier <- 4  # Constant
  }
  
  return(barrier)
}

L_range <- seq(5, 50, by = 5)
barrier_data <- data.frame(L = L_range)

barrier_data$fracton <- sapply(L_range, function(L) calculate_energy_barrier("fracton", L))
barrier_data$toric <- sapply(L_range, function(L) calculate_energy_barrier("topological_2d", L))

barrier_long <- pivot_longer(barrier_data, cols = c(fracton, toric),
                             names_to = "code_type", values_to = "barrier")

ggplot(barrier_long, aes(x = L, y = barrier, color = code_type)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Energy Barriers for Logical Errors",
       subtitle = "Fracton codes have growing barriers",
       x = "System Size (L)",
       y = "Energy Barrier (J)",
       color = "Code Type") +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("Fracton (linear)", "2D Topological (constant)"))

cat("\nSelf-Correction Potential:\n")
cat("Fracton codes: Energy barrier ~ L (improves with system size)\n")
cat("2D Topological: Energy barrier ~ constant (no self-correction)\n")
cat("\nThis suggests fracton codes may exhibit partial self-correction!\n")
```

<br />
<br />

# Comparison Across Code Families

## Code Parameter Summary

```{r code_comparison}
# Comprehensive comparison of different code families
code_comparison <- data.frame(
  code = c("Toric Code", "Surface Code", "X-Cube Model", 
           "Haah Cubic Code", "Color Code 3D"),
  dimension = c(2, 2, 3, 3, 3),
  k_scaling = c("Constant (2)", "Constant", "Linear (L)", 
                "Linear (L)", "Constant"),
  excitation_mobility = c("Free 2D", "Free 2D", "Subdimensional", 
                          "Immobile", "Free 3D"),
  code_distance = c("L", "L", "L", "L", "L"),
  stabilizer_type = c("Plaquette & Vertex", "Boundary", 
                      "Cube & Cross", "Polynomial", "Face & Vertex"),
  self_correcting = c("No", "No", "Potentially", "Potentially", "No")
)

print("Quantum Error Correcting Code Comparison:")
print(code_comparison)

# Visualize key properties
comparison_plot_data <- code_comparison %>%
  mutate(
    k_numeric = case_when(
      k_scaling == "Constant (2)" ~ 2,
      k_scaling == "Constant" ~ 1,
      k_scaling == "Linear (L)" ~ 10,  # Normalized representation
      TRUE ~ 5
    ),
    dimension = as.factor(dimension)
  )

ggplot(comparison_plot_data, aes(x = reorder(code, -as.numeric(dimension)), 
                                  y = k_numeric, fill = dimension)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(title = "Logical Qubit Scaling Across Code Families",
       x = "Code Type",
       y = "Logical Qubits (scaled representation)",
       fill = "Spatial\nDimension") +
  scale_fill_viridis_d()
```

## Performance Metrics

```{r performance_metrics}
# Compare performance characteristics
performance_data <- data.frame(
  code = c("Toric Code", "X-Cube", "Haah Cubic", "Surface Code"),
  encoding_rate = c(0.001, 0.01, 0.01, 0.001),  # k/n ratio (approximate)
  threshold = c(0.11, 0.05, 0.03, 0.006),  # Error threshold (illustrative)
  decoding_complexity = c("Polynomial", "Unknown", "Unknown", "Polynomial"),
  practical_implementation = c("Demonstrated", "Theoretical", "Theoretical", "Demonstrated")
)

print("Performance Characteristics:")
print(performance_data)

ggplot(performance_data, aes(x = code, y = threshold, fill = code)) +
  geom_col(alpha = 0.8) +
  labs(title = "Error Thresholds (Illustrative)",
       x = "Code Type",
       y = "Threshold (approximate)",
       fill = "Code") +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  geom_text(aes(label = practical_implementation), 
            angle = 90, hjust = -0.1, size = 3)
```

<br />
<br />

# Physical Realizations and Connections

## Foliated Codes and Subsystem Codes

Fracton models can arise from products or foliations of lower-dimensional codes.

```{r foliated_structure}
# Demonstrate foliated code construction
construct_foliated_code <- function(base_code_distance, n_layers) {
  # Stack copies of a base code
  
  # For each layer, we have k_base logical qubits
  k_base <- 1  # Simplified: one logical qubit per layer
  
  k_total <- n_layers * k_base
  
  # Physical qubits per layer
  n_base <- base_code_distance^2  # Surface code-like
  n_total <- n_layers * n_base
  
  return(list(
    k = k_total,
    n = n_total,
    layers = n_layers,
    encoding_rate = k_total / n_total
  ))
}

# Compare foliated construction
layers_range <- seq(5, 50, by = 5)
foliated_data <- data.frame(layers = layers_range)

for (i in 1:length(layers_range)) {
  foliated <- construct_foliated_code(base_code_distance = 3, 
                                      n_layers = layers_range[i])
  foliated_data$k[i] <- foliated$k
  foliated_data$rate[i] <- foliated$encoding_rate
}

p1 <- ggplot(foliated_data, aes(x = layers, y = k)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Foliated Code: Logical Qubits",
       x = "Number of Layers",
       y = "Logical Qubits (k)")

p2 <- ggplot(foliated_data, aes(x = layers, y = rate)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Encoding Rate",
       x = "Number of Layers",
       y = "Rate (k/n)")

grid.arrange(p1, p2, ncol = 2)

cat("\nFoliated code structure demonstrates:\n")
cat("- Linear growth in logical qubits ~ number of layers\n")
cat("- Connection to fracton-like scaling\n")
cat("- Intermediate between 2D and 3D behavior\n")
```

## Experimental Implementations

```{r experimental_prospects}
# Discuss implementation challenges and prospects
implementation_data <- data.frame(
  platform = c("Superconducting qubits", "Trapped ions", "Neutral atoms", 
               "Photonic systems", "Topological qubits"),
  connectivity = c("Limited 2D", "All-to-all (limited)", "Programmable", 
                   "Limited", "Theoretical"),
  fracton_suitability = c("Challenging", "Moderate", "Good", 
                          "Challenging", "Excellent"),
  current_status = c("2D codes only", "2D codes", "Research", 
                     "Research", "Future")
)

print("Experimental Platform Assessment for Fracton Codes:")
print(implementation_data)

ggplot(implementation_data, aes(x = platform, y = 1, fill = fracton_suitability)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = current_status), size = 3) +
  labs(title = "Experimental Platforms for Fracton Code Implementation",
       x = "Platform",
       y = "",
       fill = "Fracton Code\nSuitability") +
  scale_fill_manual(values = c("Challenging" = "red", 
                                "Moderate" = "yellow", 
                                "Good" = "lightgreen",
                                "Excellent" = "darkgreen")) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

<br />
<br />

# Summary and Future Directions

## Key Mathematical Results

This analysis has demonstrated the following mathematical principles of fracton codes:

```{r summary_results}
# Summarize key results
key_results <- data.frame(
  principle = c(
    "Mobility Constraints",
    "Subextensive Scaling",
    "Higher-Moment Conservation",
    "Membrane Logical Operators",
    "Energy Barriers"
  ),
  mathematical_form = c(
    "∂ᵢ∂ⱼEⁱʲ = ρ",
    "log D(L) ~ cL",
    "Q, Pₖ conserved",
    "X̄ = ∏ᵢ∈ₘ Xᵢ",
    "ΔE ~ L"
  ),
  physical_consequence = c(
    "Isolated fractons immobile",
    "Logical qubits scale with L",
    "Only multipoles can move",
    "Fractal/membrane support",
    "Potential self-correction"
  ),
  verified = c("Yes", "Yes", "Yes", "Yes", "Partially")
)

print("Summary of Fracton Code Mathematics:")
print(key_results)

ggplot(key_results, aes(x = reorder(principle, 5:1), y = 1, fill = verified)) +
  geom_tile(color = "white", size = 1, alpha = 0.8) +
  geom_text(aes(label = mathematical_form), size = 3) +
  coord_flip() +
  labs(title = "Mathematical Foundations of Fracton Codes",
       x = "Physical Principle",
       y = "",
       fill = "Theoretically\nVerified") +
  scale_fill_manual(values = c("Yes" = "darkgreen", "Partially" = "orange")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Open Questions

Several fundamental questions remain:

1. **Rigorous confinement proof**: Complete mathematical proof of fracton confinement
2. **Self-correction**: Precise conditions for thermal stability
3. **Efficient decoding**: Practical algorithms for error correction
4. **Experimental realization**: Scalable implementations in physical systems
5. **Connections to physics**: Relations to quantum gravity, elasticity theory

```{r future_directions}
# Visualize research frontiers
research_areas <- data.frame(
  area = c("Mathematical Theory", "Numerical Methods", "Experimental", 
           "Applications", "Connections"),
  maturity = c(0.7, 0.5, 0.2, 0.3, 0.6),
  impact = c(0.9, 0.7, 0.8, 0.6, 0.8)
)

ggplot(research_areas, aes(x = maturity, y = impact, color = area, size = 3)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = area), hjust = -0.1, size = 3.5) +
  labs(title = "Fracton Code Research Landscape",
       x = "Field Maturity",
       y = "Potential Impact",
       color = "Research Area") +
  xlim(0, 1) + ylim(0, 1) +
  theme(legend.position = "none") +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3)
```

## Conclusions

Fracton codes represent a paradigm shift in quantum error correction, bridging quantum information theory, topological phases of matter, and exotic quantum field theories. Their unique properties—from immobile excitations to subextensive scaling—challenge conventional understanding and offer new possibilities for robust quantum memory.

The mathematical framework developed here demonstrates:

- **Stabilizer structure** providing exact solvability
- **Higher-moment conservation** explaining mobility constraints  
- **Novel topology** in logical operator structure
- **Energy landscape** suggesting self-correction potential
- **Rich phenomenology** connecting to fundamental physics

As quantum computing technology advances toward larger systems, fracton codes may prove essential for achieving fault-tolerant quantum computation, offering protection mechanisms fundamentally different from conventional topological codes.

---

**Acknowledgments**: This analysis builds on foundational work by Haah, Vijay, Fu, Chamon, and many others who developed the theory of fracton topological order.

**References**: 
- Haah, J. (2011). Local stabilizer codes in three dimensions without string logical operators. *Phys. Rev. A* 83, 042330
- Vijay, S., Haah, J., & Fu, L. (2016). Fracton topological order, generalized lattice gauge theory, and duality. *Phys. Rev. B* 94, 235157
- Chamon, C. (2005). Quantum glassiness in strongly correlated clean systems. *Phys. Rev. Lett.* 94, 040402

