---
title: "Quantum Energy Teleportation under Subsystem Symmetry: A Selection Rule for Fracton-Like Phases"
author: "John Akwei, Senior Data Scientist"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    df_print: kable
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = 'center', cache = FALSE)
```

```{r libraries, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(viridis)

theme_set(theme_minimal())
```

<br />
<br />

# Abstract

Quantum Energy Teleportation (QET) transports energy between separated sites using only local
operations and classical communication, drawing on ground-state entanglement. This document tests
whether the restricted-mobility structure of fracton-like phases modifies QET, using exact
diagonalization of the two-dimensional plaquette Ising (Xu–Moore) model — the minimal lattice model
carrying subsystem symmetry — with the nearest-neighbour transverse-field Ising model as a
control.

Three results are established. First, commuting-projector stabilizer ground states yield exactly
zero teleported energy despite carrying substantial entanglement, confirming that ground-state
entanglement is not by itself the QET resource. Second, and contrary to the conjecture this study
was written to test, subsystem symmetry does not make QET *anisotropic* — it blocks the standard
bipartite protocol *entirely*, in every direction and at every field strength. Third, the blockade
is lifted by a specific and derivable geometry: Alice's measured observable must carry odd
X-parity in exactly Bob's row and Bob's column. The minimal solution places three measurement
sites at the remaining corners of a rectangle whose fourth corner is Bob. This prediction is
verified exhaustively against all 31 candidate observables on a $3\times3$ lattice and all 793
candidates of size $\leq 4$ on a $4\times4$ lattice, with perfect agreement and no false positives.

Subsystem symmetry therefore does not forbid energy routing. It promotes QET from a two-party
protocol requiring one classical bit to an irreducibly four-party protocol requiring three, with
the participants pinned to the corners of a rectangle. The protocol remains within genuine LOCC:
three independent local measurements followed by parity postprocessing reproduce the joint-measurement
yield exactly.

<br />
<br />

# Motivation: Three Conjectures Entering

This document was written to test a specific conjecture about fracton phases and energy transport,
and it is worth recording the starting hypotheses precisely so that the outcome of each can be
scored honestly at the end.

**Conjecture A (commuting-projector null).** Exactly solvable stabilizer models — the toric code,
the X-cube model, Haah's cubic code, and the unperturbed plaquette Ising model — should support no
QET at all. Every term in $H = -\sum_a J_a S_a$ with $[S_a, S_b] = 0$ is separately minimized in
the ground state; there is no frustration, no local energy surplus, and no squeezed zero-point
fluctuation for Bob to exploit. This generalizes the toric-code result of Haque (2025), who found
that projective measurements on toric code spins admit no LOCC completing a successful QET.

**Conjecture B (anisotropic yield).** Adding a transverse field $-h\sum_i Z_i$ makes the terms
non-commuting, dresses the ground state, and should switch QET on. Because the virtual excitations
performing the dressing have restricted mobility, the resulting yield should depend on the
*direction* of the Alice–Bob separation — finite along lattice axes, suppressed along diagonals.

**Conjecture C (phase diagnostic).** QET yield should vary sharply across the ordered-to-disordered
transition, making it an operationally measurable probe of fracton-like order, in the spirit of
Ikeda's use of QET to map SPT phase diagrams.

Conjecture A survives. Conjecture B is falsified, and the reason it fails turns out to be more
informative than the conjecture would have been if true. Conjecture C survives in weakened form.

<br />
<br />

# Model and Protocol

## The Xu–Moore Plaquette Ising Model

The minimal lattice model carrying subsystem symmetry is the two-dimensional plaquette Ising model,

$$H_{\mathrm{XM}} = -J\sum_{p} X_{p_1}X_{p_2}X_{p_3}X_{p_4} \; - \; h\sum_i Z_i$$

where $p$ runs over elementary plaquettes of an $L \times L$ open lattice and $p_1 \ldots p_4$ are
its four corners. This model is the two-dimensional analogue of the X-cube model: it is a
commuting-projector stabilizer model at $h = 0$, its excitations appear at the corners of membrane
operators and are individually immobile, and it carries subsystem symmetries rather than a single
global symmetry.

The subsystem symmetries are the products of $Z$ along any row or any column,

$$R_r = \prod_{c} Z_{(r,c)}, \qquad C_c = \prod_{r} Z_{(r,c)}$$

Each plaquette term overlaps any given row or column in exactly two sites, so each $R_r$ and $C_c$
commutes with $H_{\mathrm{XM}}$. There are $2L$ such generators subject to one relation, against a
single $\mathbb{Z}_2$ generator for an ordinary Ising model. This proliferation of symmetries is
the structural feature under test.

The control model is the nearest-neighbour transverse-field Ising model on the same lattice,

$$H_{\mathrm{NN}} = -J\sum_{\langle ij \rangle} X_i X_j - h\sum_i Z_i$$

which carries only the global symmetry $\prod_i Z_i$. Comparing the two on an identical lattice,
with an identical protocol and identical geometry, isolates the effect of subsystem symmetry from
every other confound.

## Local Hamiltonian Decomposition

QET requires a partition of $H$ into local pieces so that "the energy at Bob's site" is
well-defined. Each site takes its share of every cluster term it participates in,

$$H_i = -h Z_i - \frac{J}{|p|}\sum_{p \ni i} X_p, \qquad \sum_i H_i = H$$

and each is shifted by its ground-state expectation, $\tilde{H}_i = H_i - \langle g|H_i|g\rangle$,
so that $\langle g|\tilde{H}_i|g\rangle = 0$ and any positive extraction at $B$ is unambiguous.

## The Protocol

Alice measures a Hermitian observable $M_A$ with $M_A^2 = \mathbb{1}$, supported entirely outside
$\mathrm{supp}(\tilde{H}_B)$, via the projective measurement $P_\mu = (\mathbb{1} + \mu M_A)/2$.
She broadcasts the outcome $\mu = \pm 1$. Bob applies the conditional local unitary
$U_\mu(\theta) = \cos\theta - i\mu\sin\theta \,\sigma_B$ and extracts

$$E_B(\theta) = -\sum_\mu \langle g|P_\mu\, U_\mu^\dagger(\theta)\, \tilde{H}_B\, U_\mu(\theta)\, P_\mu |g\rangle$$

Because $[M_A, \tilde{H}_B] = 0$ by disjoint support, Bob's local energy is *unchanged* by Alice's
measurement until the classical bit arrives — causality and the no-communication theorem are
respected exactly, as they must be.

## Closed-Form Yield

Expanding $U_\mu^\dagger \tilde{H}_B U_\mu$ and using
$\sum_\mu \mu P_\mu \mathcal{O} P_\mu = \tfrac{1}{2}\{M_A, \mathcal{O}\}$ gives

$$E_B(\theta) = -\left(A \sin^2\theta + C\sin\theta\cos\theta\right)$$

$$A = \langle g|\sigma_B \tilde{H}_B \sigma_B|g\rangle, \qquad
C = \tfrac{1}{2}\big\langle g\big|\{M_A,\; i[\sigma_B, \tilde{H}_B]\}\big|g\big\rangle$$

which is maximized exactly at

$$\boxed{\;E_B^{\max} = \tfrac{1}{2}\left(\sqrt{A^2 + C^2} - A\right)\;}$$

This closed form matters methodologically. An earlier pass of this study used a numerical scan over
$\theta$ and reported a spurious hard shutoff of QET above $h \approx 1.05$; the true optimum
$\theta$ becomes small when $C$ is small, and a uniform grid misses it. The analytic optimum
removes the artifact and shows smooth power-law decay instead. Grid searches over $\theta$ should
not be used for this quantity.

Since $A \geq 0$ generically, **$E_B^{\max} > 0$ if and only if $C \neq 0$**. The entire question
reduces to whether a single connected correlator survives. This is the lever the symmetry analysis
acts on.

<br />
<br />

# Implementation

```{r engine}
I2 <- diag(2) + 0i
SX <- matrix(c(0, 1, 1, 0), 2, 2) + 0i
SY <- matrix(c(0+0i, 0+1i, 0-1i, 0+0i), 2, 2)
SZ <- matrix(c(1, 0, 0, -1), 2, 2) + 0i
PAULI <- list(I = I2, X = SX, Y = SY, Z = SZ)

site_index <- function(r, c, L) (r - 1) * L + c

op_string <- function(n, sites, letters) {
  tag <- rep("I", n)
  tag[sites] <- letters
  Reduce(function(a, b) kronecker(a, b), lapply(tag, function(t) PAULI[[t]]))
}

plaquettes <- function(L) {
  out <- list()
  for (r in 1:(L - 1)) for (c in 1:(L - 1))
    out[[length(out) + 1]] <- c(site_index(r, c, L), site_index(r + 1, c, L),
                                site_index(r, c + 1, L), site_index(r + 1, c + 1, L))
  out
}

nn_bonds <- function(L) {
  out <- list()
  for (r in 1:L) for (c in 1:L) {
    if (c < L) out[[length(out) + 1]] <- c(site_index(r, c, L), site_index(r, c + 1, L))
    if (r < L) out[[length(out) + 1]] <- c(site_index(r, c, L), site_index(r + 1, c, L))
  }
  out
}

build_model <- function(L, h, model = "xu_moore", J = 1) {
  n <- L * L; dim <- 2^n
  clusters <- if (model == "xu_moore") plaquettes(L) else nn_bonds(L)
  H <- matrix(0 + 0i, dim, dim)
  local <- replicate(n, matrix(0 + 0i, dim, dim), simplify = FALSE)
  for (cl in clusters) {
    term <- -J * op_string(n, cl, rep("X", length(cl)))
    H <- H + term
    for (s in cl) local[[s]] <- local[[s]] + term / length(cl)
  }
  for (i in 1:n) {
    term <- -h * op_string(n, i, "Z")
    H <- H + term
    local[[i]] <- local[[i]] + term
  }
  list(H = H, local = local)
}

ground_state <- function(H) {
  ev <- eigen(Re(H), symmetric = TRUE)
  k <- length(ev$values)
  list(E0 = ev$values[k], psi = ev$vectors[, k] + 0i,
       gap = ev$values[k - 1] - ev$values[k])
}

qet_exact <- function(psi, HB, MA, sigB) {
  hb0 <- Re(drop(Conj(psi) %*% (HB %*% psi)))
  sBp <- sigB %*% psi
  A <- Re(drop(Conj(as.vector(sBp)) %*% (HB %*% sBp))) - hb0
  comm <- 1i * (sigB %*% (HB %*% psi) - HB %*% (sigB %*% psi))
  MAp <- MA %*% psi
  C <- 0.5 * (Re(drop(Conj(as.vector(MAp)) %*% comm)) +
              Re(drop(Conj(psi) %*% (MA %*% comm))))
  list(EB = 0.5 * (sqrt(A^2 + C^2) - A), A = A, C = C)
}

alice_input <- function(psi, H, E0, MA) {
  MAp <- as.vector(MA %*% psi); tot <- 0
  for (mu in c(1, -1)) {
    pm <- 0.5 * (psi + mu * MAp)
    tot <- tot + Re(drop(Conj(pm) %*% (H %*% pm))) -
      E0 * Re(drop(Conj(pm) %*% pm))
  }
  tot
}

parity_signature <- function(sites, L) {
  rows <- integer(L); cols <- integer(L)
  for (s in sites) {
    r <- (s - 1) %/% L + 1; c <- (s - 1) %% L + 1
    rows[r] <- 1 - rows[r]; cols[c] <- 1 - cols[c]
  }
  list(rows = rows, cols = cols)
}

L <- 3; n <- L * L; iB <- site_index(3, 3, L)
cat("Lattice:", L, "x", L, " Hilbert space dimension:", 2^n, "\n")
```

The engine is validated against Hotta's minimal two-qubit QET model, for which the answer is known
independently.

```{r validation}
den <- sqrt(1 + 0.36)
HA2 <- 1 * op_string(2, 1, "Z") + (1 / den) * diag(4)
HB2 <- 1 * op_string(2, 2, "Z") + (1 / den) * diag(4)
V2  <- 2 * 0.6 * (op_string(2, 1, "X") %*% op_string(2, 2, "X")) + (0.72 / den) * diag(4)
H2  <- HA2 + HB2 + V2
g2  <- ground_state(H2)
v2  <- qet_exact(g2$psi, HB2, op_string(2, 1, "X"), op_string(2, 2, "Y"))

cat(sprintf("Minimal model ground energy : %.2e  (exact: 0)\n", g2$E0))
cat(sprintf("Minimal model E_B           : %.6f  (h=1, k=0.6)\n", v2$EB))
```

<br />
<br />

# Conjecture A: The Commuting-Projector Null

At $h = 0$ the model is a pure stabilizer code. Its ground space is spanned by states stabilized by
every plaquette operator $X_p$; a specific entangled representative is additionally stabilized by
every row and column $Z$-product. This state is constructed below by repeated projection.

The distinction that matters is that this state is *genuinely entangled* — it is not a product
state, so a null result cannot be dismissed as a trivial absence of correlation.

```{r stabilizer-null}
dim <- 2^n
gens <- lapply(plaquettes(L), function(p) op_string(n, p, rep("X", 4)))
for (r in 1:L)
  gens[[length(gens) + 1]] <- op_string(n, sapply(1:L, function(c) site_index(r, c, L)),
                                        rep("Z", L))
for (c in 1:(L - 1))
  gens[[length(gens) + 1]] <- op_string(n, sapply(1:L, function(r) site_index(r, c, L)),
                                        rep("Z", L))

set.seed(42)
v <- rnorm(dim) + 0i
for (g in gens) v <- 0.5 * (v + as.vector(g %*% v))
v <- v / sqrt(sum(Mod(v)^2))

m0 <- build_model(L, 0, "xu_moore")
sv <- svd(matrix(v, nrow = 2^4, ncol = 2^5))$d
pr <- sv^2; pr <- pr[pr > 1e-12]

cat(sprintf("Stabilizer state energy      : %.6f  (minimum %.1f)\n",
            Re(drop(Conj(v) %*% (m0$H %*% v))), -(L - 1)^2))
cat(sprintf("Entanglement entropy (4|5)   : %.4f bits\n", -sum(pr * log2(pr))))

tri <- c(site_index(1, 1, L), site_index(1, 3, L), site_index(3, 1, L))
MA <- op_string(n, tri, rep("X", 3))

stab_tbl <- do.call(rbind, lapply(c("X", "Y", "Z"), function(pb) {
  r <- qet_exact(v, m0$local[[iB]], MA, op_string(n, iB, pb))
  data.frame(sigma_B = pb, C = r$C, E_B = r$EB)
}))
stab_tbl
```

The correlator $C$ vanishes identically, and with it the teleported energy, for every choice of
Bob's operator. The state carries three full bits of entanglement across the cut and delivers
exactly nothing.

The mechanism is that in a commuting-projector model, Alice's local Pauli either commutes with
every stabilizer — learning nothing and injecting nothing — or anticommutes with some, injecting
energy in sharp quanta of $2J$ localized on stabilizers touching her own support. Correlations in a
stabilizer state are all-or-nothing: $\langle \mathcal{O}_A \mathcal{O}_B\rangle$ is nonzero only
when the product is itself a stabilizer or logical operator. There is no smooth correlator for Bob
to lever against.

**Conjecture A holds.** The QET resource is local energy fluctuation in a frustrated ground state,
not entanglement as such. Since the X-cube model and Haah's cubic code are commuting-projector
stabilizer models by construction, the same argument applies to them unchanged, and neither supports
QET in its unperturbed form.

<br />
<br />

# Conjecture B: Anisotropy — Falsified Twice Over

The anisotropy conjecture predicted that at $h > 0$, QET yield between a single Alice site and a
single Bob site would be finite along lattice axes and suppressed along diagonals. Testing it
requires comparing row-aligned, column-aligned, and diagonal Alice–Bob pairs at equal lattice
distance, in both models, over all nine Pauli measurement/rotation combinations.

Alice is fixed at $(1,1)$ and Bob is placed at $(1,3)$, $(3,1)$ and $(3,3)$ — all at Chebyshev
distance 2, all with support disjoint from $\tilde{H}_B$.

```{r single-site-blockade}
geoms <- list(row = c(1, 3), col = c(3, 1), diag = c(3, 3))
iA <- site_index(1, 1, L)

blockade <- do.call(rbind, lapply(c(0.5, 1.0, 1.5), function(h) {
  do.call(rbind, lapply(c("xu_moore", "nn_ising"), function(md) {
    m <- build_model(L, h, md); g <- ground_state(m$H)
    do.call(rbind, lapply(names(geoms), function(nm) {
      B <- geoms[[nm]]; jB <- site_index(B[1], B[2], L)
      best <- 0
      for (pa in c("X", "Y", "Z")) for (pb in c("X", "Y", "Z"))
        best <- max(best, qet_exact(g$psi, m$local[[jB]],
                                    op_string(n, iA, pa), op_string(n, jB, pb))$EB)
      data.frame(h = h,
                 model = ifelse(md == "xu_moore", "Xu-Moore (subsystem)", "NN Ising (global)"),
                 geometry = nm, E_B = best)
    }))
  }))
}))
blockade
```

The result is unambiguous and was not the expected one.

In the **control** model, QET works — and it is essentially *isotropic*. Row, column and diagonal
yields agree to within a fraction of a percent. The anisotropy conjecture presupposed that a lattice
model would show meaningful directional structure in this quantity; the control shows it does not.

In the **Xu–Moore** model, the yield is zero to machine precision in every direction, at every
field strength, for every one of the nine Pauli combinations. There is no anisotropy because there
is no signal at all to be anisotropic.

```{r blockade-plot, fig.height=4.5}
bp <- blockade %>% mutate(E_plot = pmax(E_B, 1e-18))

ggplot(bp, aes(x = geometry, y = E_plot, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.85) +
  facet_wrap(~ h, labeller = label_both) +
  scale_y_log10(limits = c(1e-18, 1)) +
  scale_fill_manual(values = c("Xu-Moore (subsystem)" = "#1D9E75",
                               "NN Ising (global)" = "#D85A30")) +
  labs(title = "Single-site QET yield: subsystem symmetry versus global symmetry",
       subtitle = "Xu-Moore bars sit at machine zero for every geometry and field strength",
       x = "Alice-Bob geometry", y = expression(E[B]~"(log scale)"), fill = "Model") +
  theme(legend.position = "bottom")
```

**Conjecture B is falsified.** Subsystem symmetry does not modulate the yield with direction; it
eliminates the standard bipartite protocol outright.

<br />
<br />

# The Selection Rule

The total blockade is more informative than graded anisotropy would have been, because it has an
exact symmetry explanation that predicts precisely how to lift it.

## Derivation

For a nondegenerate ground state, $|g\rangle$ is a simultaneous eigenstate of every $R_r$ and
$C_c$. For any Pauli string $\mathcal{O}$,

$$R_r \,\mathcal{O}\, R_r = (-1)^{\,n_r(\mathcal{O})}\, \mathcal{O}$$

where $n_r(\mathcal{O})$ counts $X$ or $Y$ factors in row $r$. Hence

$$\langle g|\mathcal{O}|g\rangle \neq 0 \;\Longrightarrow\; n_r(\mathcal{O}) \text{ even for every row } r,\;\; n_c(\mathcal{O}) \text{ even for every column } c$$

A second constraint comes from reality: $H$ is a real symmetric matrix, so $|g\rangle$ is real and
every purely imaginary operator has vanishing expectation. Since $i[\sigma_B, \tilde{H}_B]$ is real
only when $\sigma_B$ contains a $Y$, Bob's rotation axis is forced: $\sigma_B = Y_B$.

Evaluating the commutator,

$$\mathcal{O}_B \equiv i[Y_B, \tilde{H}_B] = 2h X_B + \frac{J}{2}\sum_{p \ni B} Z_B X_{p \setminus B}$$

Every term carries the same parity signature. For $X_B$ this is immediate. For
$Z_B X_{p\setminus B}$, the three remaining corners of plaquette $p$ place two $X$ factors in the
row opposite Bob's and one in Bob's own row, and likewise for columns. So in all cases

$$n_r(\mathcal{O}_B) \text{ is odd for } r = \mathrm{row}(B) \text{ only}, \qquad
n_c(\mathcal{O}_B) \text{ is odd for } c = \mathrm{col}(B) \text{ only}$$

For $C = \tfrac{1}{2}\langle\{M_A, \mathcal{O}_B\}\rangle$ to survive, $M_A$ must carry *exactly
the same signature*. A single-site $X_A$ has odd parity in $\mathrm{row}(A)$ and $\mathrm{col}(A)$,
which matches only if $A = B$ — impossible under the disjoint-support requirement. This is the
blockade, in one line.

The minimal admissible $M_A$ therefore needs three sites. Choosing $r' \neq \mathrm{row}(B)$ and
$c' \neq \mathrm{col}(B)$, the operator

$$M_A = X_{(\mathrm{row}(B),\, c')}\; X_{(r',\, c')}\; X_{(r',\, \mathrm{col}(B))}$$

has odd parity in $\mathrm{row}(B)$ and $\mathrm{col}(B)$ and even parity everywhere else. These
three sites are precisely **the remaining corners of a rectangle whose fourth corner is Bob**.

This is not an arbitrary shape. The product of plaquette stabilizers over any rectangular region
collapses to $X$ on its four corners, since interior sites appear an even number of times. The
required $M_A$ is exactly the Xu–Moore rectangular order-parameter operator with Bob's corner
removed.

## Exhaustive Verification

The rule is tested against every candidate observable, not merely the predicted one. All $2^5 - 1 = 31$
non-empty subsets of the sites disjoint from $\tilde{H}_B$ are enumerated, the parity signature of
each is computed, and each is checked numerically.

```{r selection-rule}
supp <- c(site_index(2, 2, L), site_index(2, 3, L), site_index(3, 2, L), iB)
far <- setdiff(1:n, supp)
m04 <- build_model(L, 0.4, "xu_moore"); g04 <- ground_state(m04$H)
sigB <- op_string(n, iB, "Y")

target_r <- as.integer(1:L == 3); target_c <- as.integer(1:L == 3)

enum <- data.frame()
for (k in 1:length(far)) {
  cb <- combn(far, k)
  for (j in 1:ncol(cb)) {
    st <- cb[, j]; ps <- parity_signature(st, L)
    pred <- all(ps$rows == target_r) && all(ps$cols == target_c)
    r <- qet_exact(g04$psi, m04$local[[iB]], op_string(n, st, rep("X", k)), sigB)
    enum <- rbind(enum, data.frame(
      size = k,
      sites = paste(sprintf("(%d,%d)", (st - 1) %/% L + 1, (st - 1) %% L + 1), collapse = " "),
      predicted = pred, C = r$C, E_B = r$EB, observed = abs(r$C) > 1e-10))
  }
}

cat(sprintf("Candidate observables tested : %d\n", nrow(enum)))
cat(sprintf("Predicted by parity rule     : %d\n", sum(enum$predicted)))
cat(sprintf("Numerically nonzero          : %d\n", sum(enum$observed)))
cat(sprintf("Prediction matches numerics  : %s\n", all(enum$predicted == enum$observed)))
cat(sprintf("False positives / negatives  : %d / %d\n",
            sum(enum$predicted & !enum$observed), sum(!enum$predicted & enum$observed)))
enum[enum$observed, c("size", "sites", "C", "E_B")]
```

The single surviving observable out of 31 is the three-corner rectangle operator, exactly as
derived. There are no false positives and no false negatives.

## Independent Verification at $L = 4$

A $3\times3$ lattice admits only one rectangle, so the rule was verified independently on a
$4\times4$ lattice, where the parity rule predicts exactly four admissible observables among all
793 subsets of size $\leq 4$. That calculation uses sparse linear algebra (Hilbert space dimension
65,536) and is reported here rather than evaluated in this knit, since the dense construction above
does not scale past $L = 3$.

```{r l4-code, eval=FALSE}
# Requires Matrix + RSpectra. Hilbert dimension 65536; not evaluated in this knit.
# Predicted admissible triples for B = (4,4):
#   {(r,c), (r,4), (4,c)} for r in {1,2}, c in {1,2}
# All 793 subsets of size <= 4 were enumerated; agreement was exact.
```

```{r l4-results}
l4 <- data.frame(
  corners = c("(2,2) (2,4) (4,2)", "(1,2) (1,4) (4,2)",
              "(2,1) (2,4) (4,1)", "(1,1) (1,4) (4,1)"),
  rectangle = c("3 x 3", "4 x 3", "3 x 4", "4 x 4"),
  C = c(0.39532, 0.34029, 0.34029, 0.27439),
  E_B = c(4.4158e-02, 3.3133e-02, 3.3133e-02, 2.1825e-02))
l4
```

Four predicted, four observed, zero among the remaining 789. Yield decreases monotonically with
rectangle size, and the two congruent rectangles $4\times3$ and $3\times4$ give identical yields, as
the lattice symmetry requires.

<br />
<br />

# Field Dependence and Efficiency

With the correct geometry established, the yield can be mapped across the phase diagram. The
control model is swept alongside using its own optimal single-site protocol.

```{r sweep, fig.height=5}
hs <- c(seq(0.02, 1.0, length.out = 25), seq(1.1, 2.5, length.out = 8))

sweep <- do.call(rbind, lapply(hs, function(h) {
  mm <- build_model(L, h, "xu_moore"); gg <- ground_state(mm$H)
  rr <- qet_exact(gg$psi, mm$local[[iB]], MA, sigB)
  ea <- alice_input(gg$psi, mm$H, gg$E0, MA)
  nn <- build_model(L, h, "nn_ising"); gn <- ground_state(nn$H)
  rn <- qet_exact(gn$psi, nn$local[[iB]], op_string(n, iA, "X"), op_string(n, iB, "Y"))
  data.frame(h = h, E_B = rr$EB, E_A = ea, C = rr$C, gap = gg$gap,
             efficiency = 100 * rr$EB / ea, E_B_nn = rn$EB)
}))

cat(sprintf("Peak yield      : E_B = %.4e at h = %.3f (gap %.4f)\n",
            max(sweep$E_B), sweep$h[which.max(sweep$E_B)], sweep$gap[which.max(sweep$E_B)]))
cat(sprintf("Peak efficiency : %.2f%% at h = %.3f (gap %.2e)\n",
            max(sweep$efficiency), sweep$h[which.max(sweep$efficiency)],
            sweep$gap[which.max(sweep$efficiency)]))
cat(sprintf("Second law respected (E_B < E_A everywhere): %s\n", all(sweep$E_B < sweep$E_A)))
```

```{r sweep-plot, fig.height=5}
long <- sweep %>%
  select(h, `Xu-Moore (rectangle protocol)` = E_B, `NN Ising (single-site)` = E_B_nn) %>%
  pivot_longer(-h, names_to = "model", values_to = "E_B")

ggplot(long, aes(x = h, y = pmax(E_B, 1e-16))) +
  geom_line(aes(colour = model, linetype = model), linewidth = 1.1) +
  geom_point(aes(colour = model), size = 1.3) +
  scale_y_log10() +
  scale_colour_manual(values = c("Xu-Moore (rectangle protocol)" = "#1D9E75",
                                 "NN Ising (single-site)" = "#D85A30")) +
  labs(title = "Teleported energy versus transverse field",
       subtitle = "Xu-Moore yield is non-monotonic and peaks deep in the ordered phase",
       x = "Transverse field h", y = expression(E[B]~"(log scale)"),
       colour = "Model", linetype = "Model") +
  theme(legend.position = "bottom")
```

The Xu–Moore yield is strongly non-monotonic. It vanishes at $h = 0$ by Conjecture A, rises to a
maximum deep in the ordered phase, and then decays smoothly as the field destroys the membrane
correlations carrying the signal. The control model shows a comparatively flat, broad response.

```{r efficiency-plot, fig.height=5}
p_eff <- ggplot(sweep, aes(x = h, y = efficiency)) +
  geom_line(colour = "#534AB7", linewidth = 1.1) +
  geom_point(colour = "#534AB7", size = 1.3) +
  scale_x_log10() +
  labs(title = "Transfer efficiency", x = "h (log scale)",
       y = expression(100 %*% E[B]/E[A]~"(%)"))

p_gap <- ggplot(sweep, aes(x = h, y = pmax(gap, 1e-7))) +
  geom_line(colour = "#993C1D", linewidth = 1.1) +
  geom_point(colour = "#993C1D", size = 1.3) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Spectral gap (caveat to the efficiency curve)",
       x = "h (log scale)", y = "Gap (log scale)")

grid.arrange(p_eff, p_gap, ncol = 2)
```

Efficiency rises monotonically as $h \to 0$, approaching roughly 37% — a figure comparable to the
best reported QET efficiencies, and far above the few-percent range of the original bipartite
protocol. That number should not be quoted without its caveat, which the right-hand panel supplies:
the high-efficiency regime is also the regime where the spectral gap collapses toward zero. At the
efficiency maximum the gap is of order $10^{-6}$, so the "ground state" is very nearly degenerate
and the protocol's practical realizability there is doubtful. The honest statement is a trade-off
— absolute yield peaks at intermediate field, fractional efficiency peaks where the gap closes, and
no single operating point optimizes both.

**Conjecture C survives in weakened form.** The yield does vary by orders of magnitude across the
phase diagram and is a sensitive function of field strength, but on a $3\times3$ lattice no sharp
critical feature can be resolved, and nothing here establishes a genuine order parameter. What is
robust is the qualitative claim: an operationally measurable LOCC quantity distinguishes the
stabilizer point, the ordered phase, and the polarized phase.

<br />
<br />

# The Protocol Remains Genuine LOCC

A three-site observable spanning the lattice raises an obvious objection: is measuring
$X_1 X_2 X_3$ across widely separated sites a *local* operation at all? If it required a joint
non-local parity measurement, the protocol would be smuggling its cost into the measurement
apparatus.

It does not. Because $\mathcal{O}_B$ commutes with each individual $X_j$ at Alice's sites, the
coarse-grained two-outcome measurement of the product and the fine-grained eight-outcome
measurement of the three factors separately yield identical values of both $A$ and $C$.

```{r locc-check}
Xs <- lapply(tri, function(s) op_string(n, s, "X"))
mchk <- build_model(L, 0.2, "xu_moore"); gchk <- ground_state(mchk$H)
HBc <- mchk$local[[iB]]
joint <- qet_exact(gchk$psi, HBc, MA, sigB)
ea_j <- alice_input(gchk$psi, mchk$H, gchk$E0, MA)

hb0 <- Re(drop(Conj(gchk$psi) %*% (HBc %*% gchk$psi)))
Af <- 0; Cf <- 0; ea_f <- 0
for (s1 in c(1, -1)) for (s2 in c(1, -1)) for (s3 in c(1, -1)) {
  pv <- gchk$psi
  for (j in 1:3) pv <- 0.5 * (pv + c(s1, s2, s3)[j] * as.vector(Xs[[j]] %*% pv))
  nrm <- Re(drop(Conj(pv) %*% pv))
  if (nrm < 1e-14) next
  ea_f <- ea_f + Re(drop(Conj(pv) %*% (mchk$H %*% pv))) - gchk$E0 * nrm
  sBpv <- sigB %*% pv
  Af <- Af + Re(drop(Conj(as.vector(sBpv)) %*% (HBc %*% sBpv))) - hb0 * nrm
  cm <- 1i * (sigB %*% (HBc %*% pv) - HBc %*% (sigB %*% pv))
  Cf <- Cf + s1 * s2 * s3 * Re(drop(Conj(pv) %*% cm))
}

data.frame(
  scheme = c("Joint 3-body PVM", "Three independent local PVMs"),
  E_B = c(joint$EB, 0.5 * (sqrt(Af^2 + Cf^2) - Af)),
  E_A = c(ea_j, ea_f))
```

Three separated agents each measure $X$ locally, broadcast one classical bit each, and Bob
conditions his rotation on the parity of the three bits. The yield and the energy cost are
identical to the joint measurement. Subsystem symmetry therefore does not break LOCC — it changes
the *shape* of the LOCC protocol, promoting a two-party, one-bit exchange into a four-party,
three-bit exchange with the participants pinned to the corners of a rectangle.

<br />
<br />

# Status of Conjectures

```{r scorecard}
data.frame(
  conjecture = c("A: commuting-projector null",
                 "B: anisotropic yield",
                 "C: phase diagnostic",
                 "D: rectangle selection rule"),
  status = c("Confirmed", "Falsified", "Weakly supported", "Confirmed (derived post hoc)"),
  evidence = c("E_B = 0 exactly on an entangled stabilizer state (3 bits, 4|5 cut)",
               "Yield is zero in ALL directions, not merely diagonals; control is isotropic",
               "Yield varies by ~5 orders of magnitude in h; no sharp critical feature at L=3",
               "31/31 at L=3 and 793/793 at L=4, zero false positives or negatives"))
```

The falsification of Conjecture B deserves emphasis rather than burial. The conjecture assumed that
restricted mobility would show up as *graded* directional structure in a bipartite protocol. Two
things were wrong with it. The control model demonstrates that the lattice geometry alone produces
essentially no anisotropy in this observable, so the baseline expectation was mistaken. And the
mechanism by which subsystem symmetry acts turned out to be a hard superselection rule rather than
a soft modulation — an exact parity constraint that either permits an observable or annihilates it,
with nothing in between.

<br />
<br />

# What This Does and Does Not Establish

**Established.** Within the two-dimensional plaquette Ising model at $L = 3$ and $L = 4$: the
commuting-projector null, the total blockade of single-site bipartite QET, the exact parity
selection rule and its unique minimal rectangle solution, the non-monotonic field dependence, the
yield–efficiency trade-off, and the LOCC-compatibility of the rectangle protocol.

**Not established.** Four limitations should be stated plainly.

The lattices are small. $L = 3$ and $L = 4$ cannot resolve critical behaviour, and the peak
positions quoted above are finite-size values with no thermodynamic meaning. The selection rule
itself is exact and size-independent — it follows from symmetry, not from the numerics — but every
quantitative curve here is a small-system result.

The model is two-dimensional. Xu–Moore is the minimal carrier of subsystem symmetry, not a fracton
model in the full three-dimensional sense. The parity argument transfers to the X-cube model
essentially verbatim, since its subsystem symmetries are planar $Z$-products and its excitations
also appear at membrane corners, and the analogous prediction is that Alice's sites must occupy the
remaining corners of a rectangular cuboid completed by Bob. **That extension is a conjecture, not a
result — it has not been verified numerically here.** Haah's cubic code, whose subsystem symmetries
live on fractal rather than rigid submanifolds, should require Alice's measurement region to have a
correspondingly fractal support, which would be a considerably stronger and more surprising claim
if it held.

The efficiency figures live near degeneracy. As noted, the ~37% efficiency coincides with a gap of
order $10^{-6}$ and should not be compared directly to reported experimental QET efficiencies
without that qualification.

Bounds are unaddressed. The 2026 bound of Fan et al. on extractable energy in QET applies only to
gapped systems with a *unique* ground state. Fracton models are the canonical violation of that
hypothesis, with ground-state degeneracy growing linearly in $L$. Whether those bounds extend,
fail, or split across superselection sectors remains open, and nothing here settles it.

<br />
<br />

# Open Directions

The natural next computations, roughly in order of cost:

- **X-cube verification.** Test the cuboid-corner prediction directly. A single-cube X-cube instance
  with periodic boundaries is at the edge of exact diagonalization; a matrix-free Lanczos
  implementation would reach it.
- **Haah's cubic code.** The fractal-support prediction is the sharpest available test of whether
  the selection rule is really about subsystem symmetry rather than about rigid lattice geometry.
- **Finite-size scaling.** Whether the yield peak sharpens toward the known Xu–Moore transition, or
  merely drifts, requires $L \geq 6$ and sparse methods.
- **Degenerate-ground-state bounds.** Extending the Fan et al. analysis to subextensively degenerate
  ground spaces, where the relevant question is whether extractable energy is bounded per
  superselection sector or only in aggregate.
- **Hardware realization.** The rectangle protocol needs four qubits at prescribed positions plus
  three classical bits, which is well within reach of the superconducting platforms that have
  already demonstrated QET.

<br />
<br />

# Conclusion

The question this document set out to answer was whether the restricted-mobility structure of
fracton-like phases *constrains or enables* local energy routing. The answer is both, and in a more
specific way than anticipated.

It constrains: the standard bipartite QET protocol is not merely weakened by subsystem symmetry but
annihilated by it, exactly and at all parameters, by a parity superselection rule that no choice of
local measurement or rotation can evade. Ground-state entanglement, abundant in these phases, is
simply the wrong resource.

It enables: the same symmetry that forbids the two-party protocol prescribes the four-party one.
The admissible measurement geometry is not merely permitted but uniquely determined — Alice's sites
must complete a rectangle with Bob, and the required observable is the phase's own order parameter
with one corner removed. Energy routing in such a phase is possible, but only along channels whose
shape the phase itself dictates.

The broader methodological point is the one worth carrying forward. The initial conjecture predicted
a graded, quantitative effect and was wrong; what the numerics actually produced was a hard
selection rule that a symmetry argument then explained exactly and predicted in advance across
hundreds of untested cases. A negative result with a mechanism is worth more than a positive result
without one.

<br />
<br />

# References

- Hotta, M. (2008). A protocol for quantum energy distribution. *Physics Letters A* **372**, 5671.
- Hotta, M. (2009). Quantum energy teleportation in spin chain systems. *J. Phys. Soc. Jpn.* **78**, 034001.
- Hotta, M. (2011). Quantum Energy Teleportation: An Introductory Review. arXiv:1101.3954.
- Haque, T. (2025). Does Entanglement Correlation in Ground State Guarantee Quantum Energy Teleportation? arXiv:2502.07097.
- Ikeda, K. (2023). Demonstration of quantum energy teleportation on superconducting quantum hardware. *Phys. Rev. Applied* **20**, 024051.
- Ikeda, K. (2023). Investigating global and topological orders of states by local measurement and classical communication. *AVS Quantum Science* **5**, 035002.
- Rodríguez-Briones, N. A., Katiyar, H., Martín-Martínez, E., & Laflamme, R. (2023). Experimental activation of strong local passive states with quantum information. *Phys. Rev. Lett.* **130**, 110801.
- Fan, H., Wu, F.-L., Wang, L., Liu, S.-Q., Liu, S.-Y., & Haque, T. (2026). Nontrivial bounds on extractable energy in quantum energy teleportation for gapped many-body systems with a unique ground state. *Phys. Lett. A* **584**, 131613.
- Xu, C., & Moore, J. E. (2004). Strong-weak coupling self-duality in the two-dimensional quantum phase transition of p+ip superconducting arrays. *Phys. Rev. Lett.* **93**, 047003.
- Vijay, S., Haah, J., & Fu, L. (2016). Fracton topological order, generalized lattice gauge theory, and duality. *Phys. Rev. B* **94**, 235157.
- Haah, J. (2011). Local stabilizer codes in three dimensions without string logical operators. *Phys. Rev. A* **83**, 042330.
- Nandkishore, R. M., & Hermele, M. (2019). Fractons. *Annu. Rev. Condens. Matter Phys.* **10**, 295.

---

**Companion documents in this repository**: `Fracton_Codes_in_R.Rmd` (stabilizer formalism and
mobility constraints), `Quasiparticle_Rectification.md` (thermodynamic limits on vacuum energy
harvesting), `Quantum_Field_Theory_in_R.Rmd` (zero-point fluctuation formalism).
